---
title: "Austins Data"
output: 
  html_document: 
    toc: true
    toc_float: true
    theme: cosmo
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load in and prep data

```{r}
require(ape)
raw_data<-read.csv("~/Classes/fa17/pcm/Caud/Caudata_PhyloComp_Data.csv")
cd<-raw_data[,-1]
rownames(cd)<-raw_data[,1]

ct<-read.nexus("~/Classes/fa17/pcm/Caud/caud.ultra.nexus")
require(geiger)
name.check(ct, cd)
pruning<-treedata(ct, cd)
pct<-pruning$phy
name.check(pct, cd)
head(cd)
plot(pct,show.tip.label=F,type='fan')
cd$SnoutVentLength.Mean <- as.numeric(cd$SnoutVentLength.Mean)
```


# Look for a a pair of columns to focus on

Does SVL correlate to Geographic Range?

```{r}
plot(area~SnoutVentLength.Mean,data=cd)
```

NOPE!

What about about Latitude and GeoRange?

```{r}
plot(area~lat,data=cd)
```

Meh...

Absolute Rate of Molecular Evolution vs Genome Size?

```{r}
plot(cvalue.quant~Abs.Mol.Rate,data=cd)
```

Possibly...

Elevation Area?

```{r}
plot(elevation_m.maxaltitude.quant~area,data=cd)
```

Nope.

Elevation SVL?

```{r}
plot(elevation_m.maxaltitude.quant~SnoutVentLength.Mean,data=cd)
```

Nope.


SVL MolRate?

```{r}
plot(SnoutVentLength.Mean~Abs.Mol.Rate,data=cd)
```

Nope.

SVL vs Body Width?

```{r}
plot(SnoutVentLength.Mean~log(BW_mm.bodywidth.quant),data=cd)
```

Good enough!! Let's try to do something neat.

# Standard Regression

```{r}
nonPhyloModel <- lm(SnoutVentLength.Mean~log(BW_mm.bodywidth.quant),data=cd)
summary(nonPhyloModel)
anova(nonPhyloModel)
```

# PIC

Getting rid of NA's

```{r}
getridof <- which(is.na(cd$BW_mm.bodywidth.quant))
pcd <- log(cd$BW_mm.bodywidth.quant)[-getridof]
pcd <- cbind(pcd,cd$SnoutVentLength.Mean[-getridof])
rownames(pcd) <- rownames(cd)[-getridof]
ppruning<-treedata(pct, pcd)
ppct<-ppruning$phy
name.check(ppct, pcd)
```

All better

```{r}
colnames(pcd) <- c("Width","Length")
plot(Width~Length,data=pcd)
nonPhyloModel <- lm(pcd[,1]~pcd[,2])
plot(nonPhyloModel)
summary(nonPhyloModel)
anova(nonPhyloModel)

require(phytools)
phylomorphospace(ppct,pcd,label="off",node.size=c(0.5,0.7))
width_pic <- pic(pcd[,1],ppct)
length_pic <- pic(pcd[,2],ppct)
plot(width_pic,length_pic)
shapiro.test(cbind(width_pic,length_pic))
```

Our PIC's are super non-normal, oh well!

```{r}
picModel <- lm(width_pic~length_pic+0)
summary(picModel)
plot(width_pic~length_pic+0)
abline(picModel,col="blue")
```

# PGLS

```{r}
require(nlme)
pcd <- data.frame(pcd)
pglsModel <- gls(Width~Length,
                 correlation=corBrownian(phy=ppct),data=pcd)
summary(pglsModel)
plot(Width~Length,data=pcd,pch=19,cex=0.5)
abline(nonPhyloModel,col="red")
abline(pglsModel,col="blue")
coefficients(picModel)
coefficients(pglsModel)
```

