---
title: "S2: The direct measurement of an optimal offset"
author:
  - name: Bob Week 
    url: https://bobweek.github.io
    affiliation: University of Idaho
    affiliation_url: https://ibest.uidaho.edu
  - name: Scott Nuismer
    url: https://leeef.org
date: "`r Sys.Date()`"
output:  
  distill::distill_article:
    toc: true
    toc_depth: 3
editor_options: 
  chunk_output_type: console
bibliography: bibliography.bib
---

### Measuring $\delta$

Consider the Mamre interaction data set (Pauw, personal communications) replicated in Table 1. This data summarizes the fitness effects of pollination-based interactions between individuals of the long-mouthed fly species _M. Longirostris_ and the long-tubed flower _L. anceps_. To measure the optimal offset from these data, we simply minimize the sum of the square errors between the offset-matching fitness surfaces and the fitness of each species, constraining the $\delta$'s for each surface to be the same. That is, denoting $\hat W_i^j$ and $z_i^j$ the fitness and trait value of species $i$ during observation $j$, the data can be written as

$$D=\{\hat W^i,\Delta_i\}_{i=1}^n=\{\{\hat W_1^i,\hat W_2^i\},\{z_1^i-z_2^i\}\}_{i=1}^n$$

And hence we minimize the quantity

$$L(W_1,W_2|D)\equiv\sqrt{\sum_{i=1}^n\left(\hat W_1^i-W_1(\Delta_i)\right)^2+\left(\hat W_2^i-W_2(\Delta_i)\right)^2}.$$

The fitness functions can be written as

$$W_i(z_j-z_i)=c_i\exp\left(-\frac{B_i}{2}(\delta+z_j-z_i)^2\right)$$

and hence the goal is to compute

$${\arg\min}_{c_1,c_2,B_1,B_2,\delta}(L).$$

To quantify how well our model fits the data for species $j$ we compute the coefficient of determination

$${R_j}^2\equiv1-\frac{\sum_i\left(\hat W_j^i-W_j(\Delta_i)\right)^2}{\sum_i\left(\hat W_j^i-\overline{W}_j\right)^2}, \ \ \overline{W}_j=\frac{1}{n}\sum_i\hat W_j^i.$$

These tasks can be performed numerically and _R_ code for their implementation is provided at the end of this document.

```{r, echo=F, warning=F}
require(knitr)
require(kableExtra)
require(xlsx)
mamre <- read.xlsx("Mamre interaction for Bob.xlsx",1)
mamre <- data.frame(mamre$nectar_volume_consumed,mamre$POLLEN_GRAINS_DEPOSITED,mamre$TONGUE.TUBE)
colnames(mamre) <- c("nectar consumed (mL)","pollen grains deposited","tongue length - tube length (mm)")
kable(mamre,caption="Measurements of fitness and morphological differences") %>%
  kable_styling(position = "center")
```

### _R_ code

```{r, echo=T, eval=F, layout="l-page"}
require(xlsx)
mamre <- read.xlsx("Mamre interaction for Bob.xlsx",1)
mamre <- data.frame(mamre$nectar_volume_consumed,
                    mamre$POLLEN_GRAINS_DEPOSITED,mamre$TONGUE.TUBE)
colnames(mamre) <- c("nectar consumed (mL)","pollen grains deposited",
                     "tongue length - tube length (mm)")
head(mamre)

W <- function(D,c,B,d){
  f <- c*exp(-B*(d+D)^2/2)
  return(f)
}

# collect data
W1 <- mamre$`nectar consumed`
W2 <- mamre$`pollen grains deposited`
D  <- mamre$`tongue length - tube length`
n <- dim(mamre)[1]
  
L <- function(X){
  
  # unpack parameters
  c1 <- exp(X[1])
  c2 <- exp(X[2])
  B1 <- exp(X[3])
  B2 <- exp(X[4])
  d  <- exp(X[5])
  
  l <- 0
  # accumulate errors across M. longir
  for(i in 1:n){
    l <- l + ( W(-D[i],c1,B1,d) - W1[i] )^2
  }
  
  # accumulate errors across L. anceps
  for(i in 1:n){
    l <- l + ( W(D[i],c2,B2,d) - W2[i] )^2
  }
  
  return(sqrt(l))
  
}

mins <- NULL
vals <- NULL
for(i in 1:1000){
  out <- try({optim(rnorm(5),L)})
  while(!is.vector(out)){
      out <- try({optim(rnorm(5),L)})
  }
  mins <- rbind(mins,out$par)
  vals <- c(vals,out$value)
}
i <- which.min(vals)
argmin <- exp(mins[i,])

c1hat <- argmin[1]
c2hat <- argmin[2]
B1hat <- argmin[3]
B2hat <- argmin[4]
dhat  <- argmin[5]

l1 <- 0
# accumulate errors across M. longir
for(i in 1:n){
  l1 <- l1 + ( W(-D[i],c1hat,B1hat,dhat) - W1[i] )^2/n
}
l1 <- sqrt(l1/var(W1))

l2 <- 0
# accumulate errors across L. anceps
for(i in 1:n){
  l2 <- l2 + ( W(D[i],c2hat,B2hat,dhat) - W2[i] )^2/n
}
l2 <- sqrt(l2/var(W2))

fits <- data.frame(c("M. longirostris","L. anceps"),c(c1hat,c2hat),
                   c(B1hat,B2hat),c(dhat,dhat),c(1-l1,1-l2))
colnames(fits) <- c("Species","c","B",'\u03B4','R\u00B2')

kable(fits,caption="Inferred fitness parameters") %>%
  kable_styling(position = "center")

```


