---
title: "Proposal Outline"
output: 
  html_document: 
    toc: false
    toc_float: true
    theme: cosmo
    css: style.css
---

<!--
output: 
  html_document:
    toc: true
    toc_float: true
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newcommand{\para}[1]{\left( #1 \right)}
\newcommand{\W}{\mathcal{W}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Cov}{\mathrm{Cov}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\argmax}{\mathrm{argmax}}
\newcommand{\argmin}{\mathrm{argmin}}
\newcommand{\CLAIM}{\bf{Claim}}
\newcommand{\PF}{\bf{Proof}}

## Cover Sheet

### Project Summary

Develop novel theoretical techniques to investigate the (co)evolution of ecological communities.

### Intellectual Merit

Synthesizes coevolutionary theory and community ecology.

### Broader Impacts

N/A

## Background

### Coevolution and its measurement

Coevolution has been defined as the reciprocal evolutionary response in a pair of populations to their interaction.

Here I provide the definitions of pairwise and diffuse coevolution I will be using and describe the method I've developed to measure pairwise coevolution.

  * $\mathcal{A}\dots$ abiotic selection gradient
  * $\mathcal{B}\dots$ biotic selection gradient  

With our model we get $\mathcal{A}_i=A_i(\theta_i-\bar z_i)$ and $\mathcal{B}_{ij}=\delta_{ij}\sqrt{B_{ij}}+B_{ij}(\bar z_j-\bar z_i)$, $\mathcal{B}_i=\sum_j\mathcal{B}_{ij}$, and $\beta_i=\mathcal{A}_i+\mathcal{B}_i$.

A pair of species are coevolving (with respect to the traits $z_1$ and $z_2$) if $|\mathcal{B}_{12}|,|\mathcal{B}_{21}|>0$. In light of this it seems natural to define the strength of coevolution between these species as $C_{12}\equiv G_1G_2|\mathcal{B}_{12}\mathcal{B}_{21}|$.


Sprinkle in prev research here.


In quantitative genetics the breeder's equation is the primary engine that drives many models of evolution. This equation can be written $\Delta\bar z=G\beta$ where $\Delta\bar z$ represents the change in mean trait of a population over a single generation, $G$ is the additive genetic variance of that trait and $\beta$ is the selection gradient that denotes the magnitude and direction of selection. We can partition the selection gradient into biotic $(\mathcal{B})$ and abiotic $(\mathcal{A})$ sources to obtain $\beta=\mathcal{B}+\mathcal{A}$. For a collection of species I write $\mathcal{B}_{ij}$ for the biotic selection gradient imposed on species $i$ by species $j$. The net biotic selection gradient for species $i$ is $\mathcal{B}_i=\sum_j\mathcal{B}_{ij}$. By Janzen's definition species $i$ and $j$ are coevolving iff $|\mathcal{B}_{ij}|,|\mathcal{B}_{ji}|>0$ and $G_i,G_j>0$. Note that for a pair of coevolving species it is not necessary for $|\beta_i|,|\beta_j|>0$ or for $|\mathcal{B}_i|,|\mathcal{B}_j|>0$ since the net effect of many interactions, both biotic and abiotic, may cancel with each other to produce no overall change in $\bar z_i$ or $\bar z_j$. Of course, $\Delta\bar z_i=0$ only if a delicate balance is held. By subsequently setting $\mathcal{B}_{ij}$ to zero would result in $\Delta\bar z_i=-G_i\mathcal{B}'_{ij}$ where $\mathcal{B}'_{ij}$ is the previous biotic selection gradient. Hence, the absence of change can be the response to selection implied by coevolution.

I define the __strength of coevolution__ between species $i$ and $j$ as $C_{ij}=G_iG_j|\mathcal{B}_{ij}\mathcal{B}_{ji}|$.

I define the __coevolutionary potential__ between species $i$ and $j$ as $\varrho_{ij}=B_iB_j$.
 
### Coevolution in the community context

In his talk at the "Coevolution and the ecological structure of plant-insect communities" workshop at Ohio State University in 2011, Peter Abrams discusses the expansion of the definition of coevolution to a very broad sense, including situations such as the coevolution of different morphological structures of the same species as well as the coevolution of large swaths of species within a community. Obviously my focus is on the latter, but still, this requires a broadening of the definition of coevolution layed down by Dan Janzen which has had great success in both its clarity and utility (cite nuismer 2010). Janzen's definition referred specifically to a pair of interacting populations and for this reason I advocate referring to this as pairwise coevolution, allowing the word coevolution itself to refer to the more general notion that Abrams has presented.

I define an __interaction community__ as a graph whose vertices represent species. I require this set of species to be connected in the graph-theoretic sense by edges which represent pairwise interactions. This means that for any two species, we can traverse the community via pairwise interactions to get from one species to any other species. If the graph is disconnected, I refer to this set of species as a composite community.

A __coevolving community__ is a subgraph of an interaction community that is connected in the graph-theoretic sense by coevolving pairs of species.

I make the assumption that a species is represented by one trait that mediates its interactions with any other species.

It is important to distinguish between species that have coevolved and species that are currently coevolving. The classic idea of coevolution involves the observation of a pair of traits in two interacting species that appear well matched. Textbook examples include the proboscis length of Darwins moth and the corolla length of Darwins orchid as well as the levels of TTX in the flesh of newts of the genus Taricha and the resistance to TTX in the predatory snakes Thamnophsis sertalis. These are traits that have supposedly evolved over periods of time through the corresponding interaction. What makes these cases especially distinct as putative examples of coevolution is the fact that both traits in each pair are abnormal and thus seem to have simultaneously evolved in response to the interaction. This reciprocal response to selection induced by their interaction is the definition of coevolution provided by Daniel Janzen (citation).

However, by this definition alone, these classic cases of potentially coevolved traits provide only a small subset of all the coevolution that has occurred and is occurring in nature. By this definition, species are not required to interact for long periods of time to coevolve. Instead they just need to exert reciprocal selection on one another and also respond to that selection. For any pair of interacting species there is certainly some amount of selection being imparted by one onto the other, though the magnitude of this selection may be so trivial it would escape any method of detection ever concieved. Hence, so long as there is sufficient genetic variation underlying the trait in both parties, any currently interacting pairs of species are coevolving by default. If their coevolution is having noticable impacts on the evolutionary trajectories of these species, the situation returns to the colloquial notion of coevolution.

The point is not that the currently accepted definition of coevolution needs to be modified, but that it implies a much broader set of situations than typically thought of. This observation leads neatly to notions of coevolution in the broader community context where a set of many interacting species are considered at once. Indeed, in light of this discussion the notion of an interaction community collapses to that of a coevolving community. The distinction is more psychological than biological. One notion is focused on who is interacting with whom while the other is primarily concerned with the strength by which their traits coevolve.

Returning to the point being made above, for any pair of interacing species $(i,j)$ the probaility that either $\mathcal{B}_{ij}$ or $\mathcal{B}_{ji}$ are equal to zero is zero. Using the same argument for the magnitudes of $G_i$ and $G_j$, then we will always find at least some $\epsilon$-amount of coevolution for any given pair of interacting species. This is fine; it should be understood that coevolution is continuous in value, not binary. Also, the triviality of this notion of coevolution does not remove its relevance, but emphasizes the need to measure its strength ($C_{ij}$) and significance. The distribution of $C_{ij}$ in a community can provide important information detailing the structure of the community and its potential trajectory. 

Using $C_{ij}$ we can construct other measurements that provide insight into specific coevolutionary structures of communities. For example suppose species $i$ interacts with the set $I_i$ of species and $j$ interacts with $I_j$. Set $\gamma_i=\argmax_{l\in I_i}\{C_{il}\}$. If $\gamma_i=j$ and $\gamma_j=i$ then coevolution is strongest between species $i$ and $j$ than between $i$ and any of its partners other than $j$ and likewise for $j$. As this particular strength of coevolution increases above the rest, what we think of as a pairwise coevolutionary situation emerges. However, I will go a step further and define pairwise coevolution as the situation in which $\gamma_i=j$ and $\gamma_j=i$. Whether or not pairwise coevolution is significant, that is, whether it can be detected with some method is independent of whether or not it exists. So we can have species that are technically undergoing pairwise coevolution but the intensity of pairwise coevolution may be negligible. Setting $\lambda_i=\argmax_{l\in I_i\setminus\{\gamma_i\}}\{C_{il}\}$, one potential measure of pairwise coevolution is $C^2_{ij}-C_{i\lambda_i}C_{j\lambda_j}$. But this would only make sense when $\gamma_i=j$ and $\gamma_j=i$. Instead we need a notion that describes the degree to which a species is uniquely coevolving with another species. I call this __coevolutionary affinity__ and for species $i$ it can be captured by $\alpha_i\equiv C^2_{i\gamma_i}-C_{i\lambda_i}C_{\gamma_i\lambda_{\gamma_i}}$ (note that when $\alpha_i=\alpha_{\gamma_i}$ this corresponds to a measure of pairwise coevolution since it implies that $\gamma_i=j$ and $\gamma_j=i$). The subset of coevolving pairs $(i,j)$ for which $\gamma_i=j$ and $\gamma_j=i$ are an important subset of interactions for it contains precisely those pairs that are undergoing pairwise coevolution. For theses cases I'll write $\alpha_{ij}$ instead of $\alpha_i$ or $\alpha_j$ to emphasize the pairwise relationship between species $i$ and $j$. Denoting $\Theta$ as the set of all pairs of coevolving species and $\Pi\subset\Theta$ as the set of interactions undergoing pairwise coevolution (those for which $\alpha_i=\alpha_j=\alpha_{ij}$) we can provide a cursory definition of the __pairwisivity__ of a community as

$$\pi=\frac{|\Pi|}{|\Theta|}\sum_{(i,j)\in\Pi}\alpha_{ij}.$$

This quantity is interesting for it allows us to investigate the distinction between diffuse coevolution and pairwise coevolution in the community context. When $\pi=0$ we have what I call an absolutely diffuse community. However, if there's any variability in the strength of coevolution across the community, the pair with the greatest strength of coevolution will be sure to be an element of $\Pi$. Hence, $\pi>0$ a.s. As $\pi$ increases the community tends to be more populated with pairwise coevolutionary relationships. For example, the following graph represents a community with high pairwisivity.

```{r, include=F}
require(GGally)
require(network)
require(sna)
require(scales)
# random graph
nnodes <- 20
net  <- rgraph(nnodes, mode = "graph", tprob = 1)
nett <- net
net  <- network(net, directed = FALSE)

# vertex names
network.vertex.names(net) = letters[1:10]
```

```{r, echo=F, fig.align='center', fig.width=6, fig.height=4}
nedge <- network.edgecount(net)
s <- 2
w <- exp(rnorm(nedge,log(.25)-s^2/2,s))
nlabel <- 1:nnodes
ggnet2(net, edge.size = w, node.label = nlabel)

### computing pi
first <- function(x){return(x*nnodes-x*(x+1)/2)}
mp <- function(x,y){
  y+first(x-1)-x
}
Pi  <- NULL
sg  <- NULL
cv1 <- NULL
cv2 <- NULL
wp <- w

for(j in 1:nnodes){

  biggest <- which(wp==max(wp))

  ## finding the strongest coevolving pair. this pair is gauranteed to be pairwise by definition.
  # (1,2), (1,3), (1,4), ..., (1,nnodes), (2,3), (2,4), ..., (nnodes-1,nnodes)
  # (nnodes-1) + (nnodes-2) + (nnodes-3) + ...

  i <- 1
  while(first(i)<biggest){i <- i+1}
  sp1 <- i
  sp2 <- biggest-first(i-1)+i

  t1 <- 1
  i  <- sp1+1
  coevs1 <- NULL
  while(i<nnodes){
    coevs1 <- c(coevs1,mp(sp1,i))
    if( w[biggest]<w[mp(sp1,i)] ) t1<-0
    i <- i+1
  }
  i <- sp1-1
  while(i>0){
    coevs1 <- c(coevs1,mp(i,sp1))
    if( w[biggest]<w[mp(i,sp1)] ) t1<-0
    i <- i-1
  }
  
  wc1 <- w
  wc1[biggest] <- 0
  lambda1 <- which(wc1[coevs1]==max(wc1[coevs1]))
  
  t2 <- 1
  i  <- sp2+1
  coevs2 <- NULL
  while(i<nnodes){
    coevs2 <- c(coevs2,mp(sp2,i))
    if( w[biggest]<w[mp(sp2,i)] ) t2<-0
    i <- i+1
  }
  i <- sp2-1
  while(i>0){
    coevs2 <- c(coevs2,mp(i,sp2))
    if( w[biggest]<w[mp(i,sp2)] ) t2<-0
    i <- i-1
  }
  
  wc2 <- w
  wc2[biggest] <- 0
  lambda2 <- which(wc2[coevs2]==max(wc2[coevs2]))

  if(sp1 %in% Pi) t1 <- 0
  if(sp2 %in% Pi) t2 <- 0
  
  if(t1 && t2){
    Pi  <- rbind(Pi,c(sp1,sp2))
    sg  <- c(sg,biggest)
    cv1 <- c(cv1,lambda1)
    cv2 <- c(cv2,lambda2)
  }
  
  wp[biggest] <- 0

}
pi <- (nrow(Pi)/nedge)*(sum(w[sg]^2)-sum(w[cv1]*w[cv2]))
```

The distribution of coevolution in this community is captured by the following histogram.

```{r, echo=F, fig.align='center', fig.width=6, fig.height=4}
hist(w,xlab="Coevolution",main="")
```

In this case we get $\E(C)=$ `r mean(w)`, $\Var(C)=$ `r var(w)`, and $\pi=$ `r pi`. At the opposite end of the spectrum, the following community is completely diffuse.

```{r, include=F}
# random graph
net = rgraph(nnodes, mode = "graph", tprob = 1)
net = network(net, directed = FALSE)

# vertex names
network.vertex.names(net) = letters[1:10]
```

```{r, echo=F, fig.align='center', fig.width=6, fig.height=4}
nlabel <- 1:nnodes
ggnet2(net, node.label = nlabel)
```

For this community we have $\E(C)=$ `r 0.25`, $\Var(C)=$ `r 0`, and $\pi=$ `r 0`. Lastly is an example of a community that has a moderate level of pairwisivity but would probably still be considered diffuse.


```{r, include=F}
# random graph
net = rgraph(nnodes, mode = "graph", tprob = 1)
net = network(net, directed = FALSE)

# vertex names
network.vertex.names(net) = letters[1:10]
```

```{r, echo=F, fig.align='center', fig.width=6, fig.height=4}
nedge <- network.edgecount(net)
#w <- rexp(nedge,3)
s <- .5
w <- exp(rnorm(nedge,log(.25)-s^2/2,s))
nlabel <- 1:nnodes
ggnet2(net, edge.size = w, node.label = nlabel)

### computing pi
first <- function(x){return(x*nnodes-x*(x+1)/2)}
mp <- function(x,y){
  y+first(x-1)-x
}
Pi  <- NULL
sg  <- NULL
cv1 <- NULL
cv2 <- NULL
wp <- w

for(j in 1:nnodes){

  biggest <- which(wp==max(wp))

  ## finding the strongest coevolving pair. this pair is gauranteed to be pairwise by definition.
  # (1,2), (1,3), (1,4), ..., (1,nnodes), (2,3), (2,4), ..., (nnodes-1,nnodes)
  # (nnodes-1) + (nnodes-2) + (nnodes-3) + ...

  i <- 1
  while(first(i)<biggest){i <- i+1}
  sp1 <- i
  sp2 <- biggest-first(i-1)+i

  t1 <- 1
  i  <- sp1+1
  coevs1 <- NULL
  while(i<nnodes){
    coevs1 <- c(coevs1,mp(sp1,i))
    if( w[biggest]<w[mp(sp1,i)] ) t1<-0
    i <- i+1
  }
  i <- sp1-1
  while(i>0){
    coevs1 <- c(coevs1,mp(i,sp1))
    if( w[biggest]<w[mp(i,sp1)] ) t1<-0
    i <- i-1
  }
  
  wc1 <- w
  wc1[biggest] <- 0
  lambda1 <- which(wc1[coevs1]==max(wc1[coevs1]))
  
  t2 <- 1
  i  <- sp2+1
  coevs2 <- NULL
  while(i<nnodes){
    coevs2 <- c(coevs2,mp(sp2,i))
    if( w[biggest]<w[mp(sp2,i)] ) t2<-0
    i <- i+1
  }
  i <- sp2-1
  while(i>0){
    coevs2 <- c(coevs2,mp(i,sp2))
    if( w[biggest]<w[mp(i,sp2)] ) t2<-0
    i <- i-1
  }
  
  wc2 <- w
  wc2[biggest] <- 0
  lambda2 <- which(wc2[coevs2]==max(wc2[coevs2]))

  if(sp1 %in% Pi) t1 <- 0
  if(sp2 %in% Pi) t2 <- 0
  
  if(t1 && t2){
    Pi <- rbind(Pi,c(sp1,sp2))
    sg <- c(sg,biggest)
    cv1 <- c(cv1,lambda1)
    cv2 <- c(cv2,lambda2)
  }
  
  wp[biggest] <- 0

}
pi <- (nrow(Pi)/nedge)*(sum(w[sg]^2)-sum(w[cv1]*w[cv2]))

hist(w,xlab="Coevolution",main="")
```

For this community we have $\E(C)=$ `r mean(w)`, $\Var(C)=$ `r var(w)`, and $\pi=$ `r pi`.

<!--

The intensity of pairwisivity of a community is a distinct notion from the degree distribution of the community which is more in line with the notion of specialization. Combining the degree of species in a community with the strength by which they coevolve with their partners leads to a notion I will call the __specificity__ ($\psi$) of the community. The specificity of a species is $\psi_i=\sum_{j=1}^d\alpha_{ij}$. There can be a large number of specialists in a community and yet still have $\iota=0$ if $\alpha_i\neq\alpha_j$ for all $i,j$. So specialization is 

Affinity matrix: $[\alpha_{ij}]\equiv\left[\frac{C_{ij}}{\sum_j C_{ij}}\right]$

Coevolutionary generality: $-\sum_j\alpha_{ij}\ln(\alpha_{ij})$

-->

Aside from whether or not species interact, we curious to know as to how often they interact. Given that an individual is in the midst of an interaction, we might draw the identity of the species it is interacting with from a distribution. Then it makes sense to quantify the rates of interaction by convex combinations of all members of the community. 
 

<!--

#### Inputs/Outputs formalism

One missing ingredient is the reason why certain species interact and what determines their relationship. If we focus on what kind of resources are required for each species and what kind of resources are produced by each species, we can develop a more mechanistic approach. I should save this for later as I do not have the time to develop these ideas for now.

-->

### The structure and categorization of communities

Here I describe ways of categorizing communities and refer to Shipley when discussing my focus on traits and functions.

#### Interaction types

#### Trophic structure

#### Spatial structure

### Summary

  * Coevolution: The simultaneous response to reciprocal biotic selection in a set of interacting species that are connected in the graph-theoretic sense.
  * Diffuse coevolution: When selection strengths among a set of coevolving species are of the same order.
  * Pairwise coevolution: When the selection strengths between a pair of coevolving species exceed the order of selection strengths exerted by peripheral species. 
      * Peripheral species: Non-focal coevolving species.
  * Mixed coevolution: When coevolution is neither diffuse nor pairwise.

## Research Questions

  * How common is pairwise coevolution?
  * How does succession interact with coevolution?
  * How does invasion interact with coevolution?
  * Under what conditions do specialist species evolve?

## Specific Aims and Methods

Here I describe what I'm about to describe.

#### __Aim 1: Determine the relationship between the maturity of a community and the pairwisivity of the community.__

Species don't need to interact for long periods of time to coevolve, they just need to exert significant reciprocal selection on one another (and respond to that selection). I hypothesize that in young and freshly disturbed communities there will be greater variation in the strengths of selection being imposed through biotic interactions and therefore a higher pairwisivity than through similar interactions taking place in older communties, where the frenzy of niche packing has settled down.

#### Method

I will start with the well developed theory for pairwise coevolution to derive the dynamics of the pairwisivity of the community. Since in this theory selection gradients are determined by trait differences, tracking the distribution of traits in the community will be a major component of this approach. The other major component will be based on the degree distribution. If the degree of each species (the number of other species it interacts with) tends to be low, then there is potential for high pairwisivity. Since calculating $\pi$ will not likely be analytically tractable, we will instead take a statistical approach by considering the potential values $\pi$ can take given a set of traits for the community and a topology of the interactions.

The for a community that has been freshly disturbed and time $t=0$, we might expect to see the expected pairwisivity ($\bar \pi$) take the following trajectory.

```{r, include=F}
s <- 1
nums <- seq(0,100,0.1)
pi <- runif(1,0.5,2.5)/(1+exp((nums-50)/10))
```

```{r, echo=F, fig.align='center', fig.width=6, fig.height=4}
plot(nums,pi,xlab="time",ylab="expected pairwisivity")
```

#### __Aim 2: Determine the relationship between the community type and the degree of specialization in the community.__

This aim is closely related to the above aim but has several distinct qualities. Firstly, in the above aim it is thought that pattern will hold across community types. Here we explicitly focus on different types of communities defined by the kinds of interactions present. Secondly, specialization and pairwise coevolution may appear to be similar concepts, but they are in fact quite distinct. For example, members of a community can be highly specialized but pairwise coevolution absent if there is no variance in 

#### __Hypothesis 2: Communities climax quicker with coevolution present.__

Three basic phases of succession have been described by ecologists: primary and secondary successions followed by the achievement of what has been called the climax community. The idea here is that this climax community will be reached quicker if coevolution acts to mold the traits of the commnunity members.

Extend this idea to the more general setting of community assembly instead of succession, which Scott is defining as a repeatable pattern of community assembly.

Also, hypothesis 3 can potentially be absorbed by this section.

#### Method

I plan to extract some number measuring how far along the community has evolved towards its climax community. One potential value related to this transition is NPP, though some notion of entropy may be involved as well. Studying the asymptotic nature of this value, I can compare the rates at which this value converges or increases when coevolution is present versus when coevolution is not present.

Without coevolution:

```{r, include=F}
t <- seq(0,1000,1)
NPP <- 50/(1+exp((500-t)/100))
```

```{r, echo=F, fig.align='center', fig.width=6, fig.height=4}
plot(t,NPP)
```

With coevolution:

```{r, include=F}
t <- seq(0,1000,1)
NPP <- 50/(1+exp((200-t)/75))
```

```{r, echo=F, fig.align='center', fig.width=6, fig.height=4}
plot(t,NPP)
```


Biomass tends to increase over time as communities age. We can use mean fitness of the community to track this quantity.

#### __Hypothesis 3: Invasions are more successful in the presence of coevolution for mutualistic communities but less successful in the presence of coevolution for competitive communities.__

There are two potential outcomes I focus on: 1) coevolved communities will have an increased resistance to invaders and 2) coevolved communities will be more likely to accomodate the invader through more coevolution. The reason why I choose the latter for mutualistic communities is that a new member of the community acts as a new resource to other members of the community, attracting interactions that by definition benefit it. With coevolution present, the benifits of these interactions may become amplified, increasing the probability of invasion. In contrast, I chose the former for competitive communities for it is a complementary scenario where the invader's fitness is decreased by interactions with members of the community. Since the community members have "home-field advantage" it seems likely that coevolution will act to snuff out invasions by amplifying fitness costs.

#### Method

Begin with a climax community and bombard it with immigrants and see how many take. Find a way to measure this and figure out how this measurement changes between the situations with and without coevolution.

#### __Hypothesis 4: Specialization evolves as a result of competition (and spatial structure).__

As species compete for resources, this competition can drive the evolution of specialization.

#### Method

Use the inputs/outputs formalism to derive metrics of specialization/generalization and a distribution of these quantities across the community.

Find a distribution of complexity.

## Significance of Proposed Research

The proposed work will develop a set of theoretical approaches for understanding the evolution and potential coevolution of ecological communities. The goal is to synthesize ideas and theory in order to produce novel insights and a deepend understanding of the processes that shape the patterns we see in nature. This work can potentially be extended to include more idiosyncratic details that ties theory to data upon which methods of inference can be developed to assertain an explicit understanding of the empirical world. 

<!-- In a time of great flux, it is necessary that we work towards an understanding of the relationship between species and their environment (both abiotic and biotic) as well as our relationship to these species so we can best predict the outcomes of human development and climate change for the sake of the safety of future generations. -->

## Broader Impacts

I have none to declare.

## References






