---
output:
  html_document:
    css: style.css
    theme: cosmo
    toc: no
    toc_float: yes
title: "A coevolutionary theory of mutualistic communities"
author: "Bob Week"
csl: evolution.csl
bibliography: ref.bib
---

```{r setup, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)
```

```{r results="asis", echo=F}
cat("
<style>
a {
   color: black;
}
</style>
")
```

\newcommand{\W}{\mathcal{W}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\V}{\mathbb{V}}
\newcommand{\para}[1]{\left( #1 \right)}
\newcommand{\sqbr}[1]{\left[ #1 \right]}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Cov}{\mathrm{Cov}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\Matrix}[1]{\left(\begin{matrix}#1\end{matrix}\right)}

## Introduction to proposed work

Coevolution, the reciprocal evolutionary response of two populations to their interaction, has long been thought to play a key role in driving the exaggeration of traits [@darwin1859origin;@ehrlich1964butterflies], promoting major evolutionary transitions such as the evolution of sexual reproduction [@hamilton1980sex;@jaenike1978hypothesis;@lively1987evidence;@gandon2007evolution], manipulating the structure of ecological communities [@nuismer2013coevolution], influencing epidemiological dynamics [@anderson1982coevolution;@lion2015evolution;@lively2016coevolutionary], and maintaining the diversity of life (Yoder and Nuismer 2010). However, despite coevolution's long suspected importance, a rigorous understanding of both the mechanisms involved and the patterns produced by this process is still in its infancy. A clear understanding of the role that coevolution plays in the evolutionary process must account for the details of the mechanisms driving coevolution, be able to quantify its strength and significance in nature and determine its consequences for ecological communities. The proposed research seeks to develop innovative and novel contributions aimed at accomplishing these three main objectives that push beyond previous theoretical work.

Biological communities are notoriously difficult to capture mathematically because they consist of a dizzying magnitude of complexity. This complexity is manifested both in the network structure of the ecological interactions they are composed of as well as the processes that emerge from these interactions (such as nutrient cycling and community assembly). Since every species belongs to a community, confronting this complexity is a necessary task not only for community and ecosystem ecologists, but also for evolutionary biologists studying a single species. A cursory peek into past work reveals that we can incorporate a mind-numbing amount of detail involving the mating structure, genetic architecture, spatial dynamics and so on into models of the evolution and ecology of a species. But even with such a thorough investigation we typically cannot claim to have understood why this particular species is the way that it is because species do not exist in a vacuum. They co-occur with various other species and this geographic overlap inevitably leads to interspecific interactions. Of course there are models that have been developed to capture the evolutionary and ecological dynamics of interacting species (cite roughgarden, abrams etc). However, these models are typically limited to a small number of species even though we know that communities do not generally consist of just a few species. Incorporating such facts into our methods can alter the conclusions drawn. To be certain that, for example, the abiotic environment is responsible for the diversity of a species or clade, this hypothesis should be tested against competing ones such as the hypothesis that the observed diversity is the result of interspecific interactions. 

Here the focus is not on diversity measured in terms of species richness nor is it on macroevolutionary processes. This work focuses on the microevolution and diversity of functional traits across a community derived from the ecological interactions occurring within that community. The basic aim is to develop a body of theory capable of testing biotic hypotheses for community-level phenomena. For example, is the distribution of generalization/specialization across a community the result of the form of interactions taking place, or is it due to environmental factors such as spatial or temporal vicariance?

To develop such a theory, various technical obstacles must be confronted. A major limiting factor of past theoretical work comes from a common modelling approach. Previous models capturing the relationship between community ecology and coevolution have focused on results pertaining to a small number of interacting species (cite roughgarden, abrams, holt and mcpeek for theory, strauss, benkman, berenbaum, zergall for empirical work). The common approach in each of these studies is to derive a model of mean trait evolution alongside a model of demographic dynamics for each species involved. Hence, if there are N species under consideration, there will be 2N dynamical, often nonlinear, equations used to model the system. This approach has provided some analytical results for a few species, but rapidly becomes intractable with additional species and becomes heavily dependent on idiosyncratic details, derailing any hope of elucidating general principles. However, many of the patterns observed in ecological communities span dozens, even hundreds of species [@morin2009community;@hubbell2011unified;@manning1997themoegistorhynchus]. To overcome this difficulty a recent development in coevolutionary theory has taken the so-called mean-field approach with respect to the identity of species (cite paper that's in review and scott's 2013 paper and a book or article on the mean-field approach). This approach allows the reduction of a high-dimensional problem involving equations describing the evolution of each species within a community to a low-dimensional one that focuses on the community as a whole. With this approach analytical results can be derived that capture the statistical nature of the community, but without explicit conclusions for any particular member of the community. Hence, the stage has been set to develop a mathematically rigorous theory of coevolutionary community ecology. The ultimate goal of the proposed work is to contribute to this development through extending previous work on the coevolution of mutualistic networks by incorporating further biological elements than have been previously considered (such as behaviour and demographic dynamics, see section Q3).

However, before such a contribution can be made there are two major steps to be taken. First, a more biologically realistic treatment of the mechanisms mediating the outcomes of ecological interactions needs to be incorporated into coevolutionary theory. The term mechanism here refers to the set of behavioral, physiological and morphological elements dictating the outcome of an interaction. For example, the length of the proboscis of a pollinator and the depth of the nectar tube of a flower have been shown to be very important in regulating the amount of nectar obtained by the pollinator and the number of pollen grains deposited by the flower [@pauw2009flies]. Variation in this mechanism alone may explain a significant amount of functional diversity seen in nature. Thus, the first chapter will focus on deriving a new model of coevolution from first principles that takes into account more biological detail of the mechanism through which species interact. This new model will then be compared to an empirical data set involving a fly/flower pollination mutualism in order to demonstrate the amount of diversity that can explained purely by the structure of the interaction. The primary question driving this chapter is; Can the structure of an interspecific interaction explain the patterns observed in the exaggeration of functional traits?

Secondly, a metric for the strength of coevolution within a community is required to quantify the role of coevolution in the community. A way to measure the strength of coevolution between a pair of species will be needed in order to derive this community-wide metric from first principles. Hence, a novel metric of pairwise-coevolution will be introduced. Because this metric provides an opportunity to tie theory to data, an important bridge to cross whenever possible, a novel method to estimate pairwise-coevolution will be the topic of the second chapter. The question motivating this chapter is; How can we measure the strength of coevolution in the wild?

Accomplishing these first two steps will pave the way to brandishing a framework ripe for rigorous investigations of the relationship between coevolution and ecological communities. The goal of the third and final chapter will be to use the new coevolutionary model developed in the first chapter to construct a model that tracks the statistical distribution of species-level mean traits, the distribution of the intensity of pairwise-coevolution as introduced in chapter two, and the distribution of specialization/generalization of species across the community. The underlying question this chapter focuses on is; How does the structure of interactions between two guilds of species in a community affect the distribution of specialization/generalization?

To tether this work to the empirical world, each step is motivated by the details of specific pollination interactions. The primary example used throughout is the interaction between a long-tongued fly Moegistorhynchus longirostris and the flower it pollinates, Lapeirousia anceps, occuring in the coastal lowlands of South Africa [@pauw2009flies; @manning1997themoegistorhynchus]. Other fly-flower mutualisms from across the globe will be explored for cross-system comparisons.

## Q1: Can the structure of an interspecific interaction explain the patterns observed in the exaggeration of functional traits?

#### (Introduction to the offset-matching model of coevolution)

I begin by discussing the details of the mechanism mediating many pollination interactions and its implications for trait evolution. Historical characters are used to provide context, but the focus here is on the biology of interspecific interactions.

Mutualistic interactions provide an abundant supply of putative coevolutionary arms races. The most popular of these being the potentially coevolved orchid Angraecum sesquipedale and the moth that pollinates it Xanthopan morganii praedicta. The extreme length of this orchid’s nectar tube captured the attention of Charles Darwin [-@darwinvarious]. Darwin formulated the hypothesis that this exaggerated trait may be the result of the reciprocal evolutionary responses between this orchid and whatever had been pollinating it (unknown to Darwin at the time) over many generations. In a rebuttal to a criticism of Darwins idea, Alfred Russel Wallace provided some logic to defend Darwin [@arditti2012good]. Wallace had been carefully thinking about the mechanism mediating the outcome of this interaction. He speculated that the orchid would be better off having a nectar tube that is slightly longer than the proboscis of the pollinator, forcing the pollinator to further penetrate the flower in order to obtain its nectar reward. Wallace speculated that this would cause the body of the pollinator to rub against the reproductive organs of the flower and thereby increase the probability of pollination. Hence, it can expected that selection induced by this interaction will select for slightly longer tube lengths. Likewise, with respect to the fitness of the pollinator, Wallace postulated that by having a proboscis slightly longer than the tube length of the flower, the pollinator would be able to extract more resources from each flower it visits. Thus, selection from this interaction would be reciprocally selecting for individuals from both populations with slightly longer traits. Assuming that these traits maintain ample genetic variation and that the primary source of selection on these traits is induced by their interaction, then, ignoring other processes such as migration and drift, the logical result is a never ending arms race towards more and more exaggerated, yet closely matched, traits.

Given the logic of Wallace’s argument, one might expect any pollination-based mutualism mediated by a mechanism analogous to the one described above to consistently produce highly exaggerated traits. However, a wide variety in the degree of exaggeration among traits mediating such interactions across a variety of plant-pollinator interactions has been observed [@anderson2010predictable]. What can account for such variance? Clearly there is an endless list of potential explanations. However, by focusing on explanations pitched entirely in terms of the structure of the interactions between pairs of species, a clear understanding of the role that coevolution plays in the maintenance of diversity can be developed.

Since different species are often placed under different physiological constraints and environmental conditions, it seems obvious that this variation should exist across species. However, even within species, a striking amount of variation in the outcomes of these processes has been observed. Take, for example, the interaction between the fly Moegistorhynchus longirostris and the flower it pollinates, Lapeirousia anceps [@pauw2009flies]. The mechanism mediating this interaction is precisely the same as that described above for A. sesquipedale and X. morganii (see Figure X). The mean traits of the fly and flower species covary spatially with a correlation coefficient of 0.78, seeming to confirm the prediction that these traits should be closely matched. Yet, across eight populations the average length of the fly's proboscis ranges from 43 mm to 86 mm and the average nectar tube of the flower across the same populations from 41 mm to 77 mm. Then, given that each population is undergoing the same interaction, why is there such stark phenotypic variability across these species' distributions? To answer this question I propose to use a novel quantitative genetic model of coevolution derived from considering the mechanism mediating pollination interactions.

→ Image of fly and flower parts ← (figure X)

### Derivation of the model:

Previous work on the coevolutionary theory of quantitative traits have focused on two primary mechanisms; trait matching and trait differences. In the case of mutualistic trait matching, it is assumed that the optimal phenotype of each species is the mean phenotype of the complementary species [@kiester1984models]. These models track the change in mean phenotype of each species. They can be derived from “first principles” by considering the fitness of an individual as a function of its trait value along with the trait value of its interacting partner. For the mathematics to be tractable it is required that fitness as a function of the trait takes the form of a Gaussian function. Assuming each individual interacts with precisely one individual of the complementary species, fitness will be of the form

(@fit1) $$W_i(z_i,z_j)=W_{i,\max}\exp\para{-\frac{(z_i-z_j)^2}{2\beta_i^2}}$$

where $z_i$ is the trait of an individual from species $i$ and $z_j$ is the trait of an individual of species $j$. The greek letter $\beta_i^2$ is the "width" of biotic selection being imposed on species $i$ from species $j$. By further assuming the distributions of traits in each species, denoted $\phi_i(z_i)$ and $\phi_j(z_j)$ respectively, are normal we can calculate the mean fitness of each species. 

(@meanfit1) $$\overline W_i=\int_{\R^2} W_i(z_1,z_2)\phi_1(z_1)\phi_2(z_2)dz_1dz_2$$

$$=C\exp\para{-\frac{(\bar z_i-\bar z_j)^2}{2(\beta_i^2+\sigma_1^2+\sigma_2^2)}}$$

where $C$ is a compound constant involving the phenotypic variances.

After a weak selection approximation ($\beta_i^2\gg\sigma_1^2+\sigma_2^2$), these mean fitnesses can be plugged into the Breeder’s equation ($\Delta\bar z_i=G_i\partial_{\bar z}\ln\overline W_i$) to obtain a model of mean trait dynamics. This model results in very well behaved linear dynamics described by

(@match) $$\Delta\bar z_i=G_iB_i(\bar z_j-\bar z_i).$$

Here, $G_i$ denotes the additive genetic variance of the trait $z_i$, the small non-negative parameter $1≫Bi_≥0 $is the strength of biotic selection induced by species $j$ onto species $i$ and $z_i,z_j$ represent the mean traits in species $i$ and $j$ respectively. This model predicts the mean traits for each species to converge to a common value that lies somewhere between their initial phenotypes. Hence, no escalation can occur in this model.

In contrast, Nuismer et al. [-@nuismer2010correlation] developed a model based on trait differences that leads to pure escalation dynamics. One way to derive this model is to assume fitness follows a Gaussian that is translated a distance equal to the "root-width" of biotic selection. Here the width of biotic selection is denoted by $\beta_i^2$ and its root is $\beta_i$. See Lande [-@lande1976natural] for a brief discussion on this quantity. Hence,

(@fit2) $$W_i(z_i,z_j)=W_{i,\max}\exp\para{\frac{(z_j+\beta_i-z_i)^2}{2\beta_i^2}}.$$

Under the assumption of weak selection this distance is quite large and the resulting dynamics are approximated by

(@diff1) $$Δ\bar z_i=G_i\sqrt{B_i} + G_i B_i (\bar z_j-\bar z_i).$$

Alternatively, this model can be derived by modelling fitness with the sigmoid function (Figure (@sigmoid))

(@fit3) $$W_i(z_i,z_j)=\frac{W_{i,\max}}{1+e^{B_i(z_j-z_i)}}$$

```{r, include=F}
s <- seq(from=-100,to=100,by=.1)

# SIGMOIDD!!!
d<-0.025

S <- 1-1/(1+exp(2*d*s))

yl <- expression('W'['i']('z'['i'],'z'['j']))
xl <- expression('z'['i']-'z'['j'])
```

```{r sigmoid, echo=F, fig.cap="(@sigmoid) \\ Sigmoid Fitness Curve", fig.align='center', anchor="figure", fig.width=6, fig.height=4}
plot(s,S,cex=0.01,xlab=xl,ylab=yl)
```

Unfortunately this function is not friendly towards the clean algebraic analysis that Gaussian functions comply with and thus requires some extra steps. The extra steps involved are a series of Taylor approximations around $B_i$ for resolving the nonlinearities that arise from taking this route. At each Taylor expansion, the approximation involves keeping terms that are linear with respect to $B_i$, but neglects higher order terms. This results in the exact same dynamics as in eqn (@diff1).

Originally Nuismer et al. [-@nuismer2010correlation] took these approximations a step further by assuming terms of order $B_i$ and higher are negligible. This resulted in the pure escalation model resulting from trait differences expressed as

(@diff2) $$\Delta\bar z_i=G_i\sqrt{B_i}$$

Under this model the mean traits of each species will forever increase at a constant rate independently of one another. Though this model predicts an arms race, it misses a crucial ingredient described by Wallace, that the traits in each species need only be slightly longer than the traits of the complementary species to maximize fitness.

To obtain a model that captures Wallace’s notion of offset, I considered the fitness function

(@fit4) $$W_i(z_i,z_j)=W_{i,\max}\exp\para{-\frac{(z_j+\delta_i-z_i)^2}{2\beta_i^2}}$$

where the additional term $\delta_i$ represents the offset between the traits of species $i$ and $j$ that maximizes the fitness benefits of species $i$ with respect to their interaction.

Using the same series of steps described for the derivation of the trait-matching model, the trait dynamics result in

(@offmatch) $$Δ\bar z_i=G_iB_iδ_i+G_iB_i(\bar z_j−\bar z_i).$$

By setting $δ_i=0$ for both species the classic trait matching model, eqn (@match), is retrieved. In hindsight of the above discussion it is easily seen that by setting $δ_i=1/\sqrt{B_i}$, the offset-matching model is approximated by the pure escalation model when $|\bar z_j−\bar z_i|$ is small. Hence, this new model, which has been dubbed the offset-matching model of coevolution, has coalesced two previously distinct models of coevolution into one unifying framework.

### Outline of the approach:

Using the offset-matching model of coevolution, both the escalation and spatial covariation observed in nature can be explained. This offset-matching model can be used as a platform to understand why there is such variation in the outcomes of these arms races across the spatial distribution of coevolving pairs by focusing on the mechanism of the interaction as manifested by the offset parameter $δ$. 

Returning to pollination as a motivating example the optimal offset will vary for each individual and each interaction an individual participates in. The offset for a fly will depend on, for example, the behavior of the fly (such as its flight pattern if it hovers while feeding) and the amount of nectar in the tube of the flower. Thus, by optimal offset for an individual it is meant the offset that will on average benefit this individual the most across any potential interaction. This offset then clearly has a variance within a population and, assuming this variance has a genetic component, is subject to evolution by natural selection. In this initial investigation, however, I will ignore selection on the optimal offset in both plant and pollinator, but will allow this offset to evolve neutrally (by drift, mutation and migration).

Further examination of the model reveals that no equilibrium exists. Hence, in order to impose an equilibrium a source of stability, such as stabilizing selection, must be incorporated. Typically this is referred to as abiotic selection and represents the net selective pressure occurring outside of this interaction. But this adds further complexity to the model, creating dependencies on the strength of abiotic selection and the values of the abiotic optima, and distracts from the focus on an explanation pitched entirely in terms of the interaction. To avoid this unnecessary and unfavorable complication, an alternative approach to the analysis of the model relaxes the assumption of equilibrium and instead assumes these species have been interacting for some finite amount of time $T$. The parameter $T$ is relevant to a biotic explanation and hence brings the focus back to the interaction.

It will be assumed that these two species originally co-occurred in one location before diffusing across the landscape and that the strengths of biotic selection ($B_1$ and $B_2$) remained constant in space and time. This points to three sources for the variability observed; 1) the stochastic effects of drift, 2) spatial and temporal variability in the offset term and 3) the duration of the interaction $T$. None of these sources of variation are mutually exclusive. Indeed 1) depends on 3). Thus some combination of them will be employed in attempt to explain observed patterns.

Aside from $T$, the key parameters to be investigated are the variance of drift $(V_d)$ and the variance in offset $(V_δ)$ for each species. In order to couple this theory to the empirical world I plan to use an evolutionary algorithm to find values for $(V_d,V_δ,T)$ that produces simulated data whose summary statistics roughly match those of the data published in Pauw et al. [-@pauw2009flies]. For this algorithm to work, a “population” of values for the triple $(V_d,V_δ,T)$ is provided upon which a subset is selected for "reproduction." The criteria for selection will involve the comparison of summary statistics calculated from data simulated under these parameter pairs to the summary statistics drawn from the empirical system. Reproduction involves imperfect replication of the selected pairs, allowing the algorithm to search parameter space. This approach is not as formal as maximum likelihood or Bayesian approaches and is similar in many ways to Approximate Bayesian Computation. However, the goal of this project is not to do formal inference, but to show the utility of this model for describing natural phenomena. A true implementation for inference requires a depth of work that is not feasible in the time allotted for this project.

Finally, though the quantitative genetic approach has the strength of providing analytic results, its weakness lies in the large amount of assumptions required to obtain such results. To account for this restriction, I plan to test the depency of my conclusions on the assumptions made by developing genetically-explicit individual-based simulations that allow me to numerically derive results for generalized models. By comparing analytical results to simulations, I can quantify the extent to which particular assumptions prevent accurate descriptions of biological processes. Specifically, statistical tests, such as the t-test, can be used to calculate whether or not simulated results significantly differ from those predicted by quantitative genetic theory.

## Q2: How can we measure the strength of coevolution in the wild?

In spite of the much anticipated importance of the coevolutionary process [@thompson1994coevolutionary], there is virtually zero quantitative knowledge of its strength and prevalence in the wild. This lack of knowledge is largely due to the absence of generally applicable statistical approaches that yield quantitative estimates for the strength and significance of coevolution. 

Our current understanding of coevolution's importance rests upon methods that fall into two general classes: those that are broadly applicable but yield only qualitative evidence for coevolution and those that produce quantitative estimates for the strength of coevolution but can be applied only in a narrow range of systems. For example, one popular approach for inferring coevolution relies on measuring the spatial correlation between traits of interacting species and using significant interspecific correlations as evidence of a coevolutionary process [@berenbaum1986constraints; @benkman2003reciprocal; @hanifin2008phenotypic; @toju2008fine; @pauw2009flies]. Strengths of this approach include the relative ease of collecting the relevant data and its broad applicability to a wide range of species interactions. The critical weakness of this approach, however, is that significant interspecific correlations are neither necessary nor sufficient for demonstrating coevolution [@nuismer2010correlation; @janzen1980coevolution] and can never yield more than qualitative evidence for coevolution. 

In contrast, quantitative approaches for estimating the strength of coevolution, such as selective source analysis, require the collection of extensive trait and fitness data from interacting species and thus have proven difficult to employ in all but a few specialized systems [@ridenhour2005identification; @brodie2003reciprocal; @sletvold2010pollinator; @toju2005imbalance]. As a consequence of these trade-offs in existing approaches, rigorous, quantitative estimates of the strength of coevolution in natural populations are extremely scarce.

To fill this critical gap I propose a novel statistical method of inference that derives maximum likelihood estimates for the strength of pairwise coevolution in natural systems by coupling the offset-matching model of coevolution to easily collected phenotypic data. I will apply this method to two well-studied interactions: the highly specific antagonism between the Japanese camellia fruit and its predatory weevil and the mutualism between the long-tongued fly M. longirostris and the flower it pollinates L. anceps. Broad application of this approach has the potential to answer long-standing evolutionary debates such as the importance species interactions play in the evolution of sexual reproduction [@hamilton1980sex].

### Outline of the approach:

The offset-matching model can be easily extended to predict the spatial distribution of population-level mean traits for a pair of interacting species that evolve in response to random genetic drift, abiotic selection, and coevolution. This has been done by Nuismer et al. [-@nuismer2010correlation] for both the trait-matching and pure-escalation models. Crucially, this extension predicts that the distribution of trait means in the interacting species across a metapopulation will approach a bivariate normal distribution entirely described by five statistical moments: the metapopulation-level average of the population-level average key trait in each species among populations ($μ_1$ and $μ_2$), the variance of the population-level mean trait in each species among populations ($V_1$ and $V_2$), and the spatial association (covariance, $C$) between the key traits in each species. The key model parameters that quantify the strength of biotic selection acting on the two species are $B_1$ and $B_2$. Any selective agents acting independently of the interaction are captured collectively by the strengths of abiotic selection on each species; $A_1$ and $A_2$. All four of these model parameters are required to be non-negative and small ($≪1$). Denoting the optimal phenotype with respect to abiotic selection by θi and the effective population size by ni for species i, classical quantitative genetic assumptions lead to the following recursion equations for the statistical moments describing the metapopulation:

(@dynmet) $$\mu_i'=\mu_i+G_i\sqbr{B_i\delta_i+B_i(\mu_j-\mu_i)+A_i(\theta_i-\mu_i)}$$

$$V_i'=(1-2A_iG_i)V_i+2B_jG_j(C-V_i)+G_i/n_i$$ 

$$C'=(1-A_1G_1-A_2G_2)C+B_2G_2(V_1-C)+B_1G_1(V_2-C).$$

Because the models predict a bivariate normal distribution of traits, calculating the likelihood of observing any particular set of trait values in a pair of interacting species is straightforward. In particular, letting $D=\{D_i\}_{i=1}^N\equiv\{(\bar z_1,\bar z_2)_i^\top\}_{i=1}^N$ denote the data (mean trait pairs of the interacting species from several populations) one obtains

(@lik) $$L(D|\mu,\Sigma)=\prod_{i=1}^N\frac{\exp\para{-\frac{1}{2}(D_i-\mu)^\top\Sigma^{-1}(D_i-\mu)}}{2\pi\sqrt{|\Sigma|}}, \ \Sigma=\Matrix{V_1 & C \\ C & V_2}.$$

Using the equilibrium of system (4) expressions of the first five moments can be derived in terms of the selection strengths $A_1,A_2,B_1$ and $B_2$. Unfortunately these functions cannot be inverted to obtain the selection strengths as functions of the first five moments. Thus, I numerically solve for the $A_i$ and $B_i$ that maximize

(@likstr) $$L(D|A_1,A_2,B_1,B_2)\equiv L(D|\mu(A_1,A_2,B_1,B_2),\Sigma(A_1,A_2,B_1,B_2)).$$

By maximizing this likelihood with respect to the strengths of biotic and abiotic selection acting on the two species a quantitative estimate for the strength of coevolution is obtained under the assumption that the remaining model parameters have been accurately estimated.
In addition to providing quantitative estimates for the strength of biotic selection, this method can be used to rigorously test for the presence of coevolution. Specifically, for a coevolutionary hypothesis to be supported, reciprocal selection must be demonstrated (Janzen 1980; Thompson 1994). In this maximum likelihood framework, the long-standing and widely accepted definition of coevolution corresponds to demonstrating that the pair of parameters in this model that quantify the strength of biotic selection imposed on each species by the other are significantly greater than zero. By performing likelihood ratio tests, support for the coevolutionary hypothesis can be compared relative to support for the non-coevolutionary null hypotheses of unilateral evolution where $B_1=0$ or $B_2=0$, or evolution entirely independent of species interactions where $B_1=0$ and $B_2=0$ (Figure @networkfig). Due to the nested structure of these models, I can directly compare the likelihood of coevolution against the likelihoods of these null models by approximating p-values. Figure @networkfig shows that both p1 and p2 need to be less than the significance threshold α to support a coevolutionary hypothesis.

```{r, include=F}
require('jpeg')
plot_jpeg = function(path, scale, add=FALSE)
{
  jpg = readJPEG(path, native=T) # read the file
  res = scale*dim(jpg)[2:1] # get the resolution, [x, y]
  if (!add) # initialize an empty plot area if add==FALSE
    plot(1,1,xlim=c(1,res[1]),ylim=c(1,res[2]),asp=1,type='n',xaxs='i',yaxs='i',xaxt='n',yaxt='n',xlab='',ylab='',bty='n')
  rasterImage(jpg,1,1,res[1],res[2])
}
```

```{r networkfig, echo=F, fig.cap="(@networkfig) \\ Relationships between the coevolutionary hypothesis and non-coevolutionary hypotheses", fig.align='center', anchor="figure"}
par(mar=c(0,0,0,0))
plot_jpeg('~/Documents/Website/bobweek.github.io/network.jpg',2)
```

Restricting the selection parameters $B_i$ to match the null hypotheses of unilateral evolution ($B_1=0$ or $B_2=0$) results in a restricted likelihood ($L(D|A_1,A_2,0,B_2)$ or $L(D|A_1,A_2,B_1,0)$), one that is always less than $L(D|A_1,A_2,B_1,B_2)$. Then calculating the significance of $B_1$, for example, is equivalent to computing the probability that the statistic

(@likrat) $$\Lambda=2\ln\para{\frac{L(D|A_1,A_2,B_1,B_2)}{L(D|A_1,A_2,0,B_2)}}$$

lands within a given percentile of its distribution under the assumption of the null hypothesis $B_1=0$. I adopt the standard 95th percentile as the threshold for determining significance.

According to Wilks' Theorem, the distribution of this statistic is approximated by a Chi-square for large sample sizes. However, typical numbers of populations sampled are on the order of ten. Hence, this approximation fails and empirical distributions need to be calculated  [@wasserman2013all]. To do so, 250 sets of data under the null model are simulated for each comparison, with each set of data containing the number of populations equal to those sampled empirically. This provides adequate resolution (p-value estimates span 0 to 1 in increments of 0.004) while remaining computationally feasible. Calculations have been carried out in the statistical programming language $R$ and scripts will be made available at:

https://github.com/bobweek/inferring.coevolution

### Current Results:

I applied this maximum likelihood approach to two textbook examples of coevolution where previous work has implicated coevolution as a cause of trait exaggeration and spatial variability [@pauw2009flies; @toju2011weevils]: the interaction between the long-tongued fly M. longirostris and a plant it pollinates L. anceps and the interaction between the camellia plant Camellia japonica and its seed predator, the weevil Curculio camellia. In both cases, the interactions are thought to depend largely on a single key trait in each species (fly proboscis and plant floral tube lengths or weevil rostrum length and camellia pericarp thickness). In addition to the essential phenotypic data, previous work in both systems provided valuable additional information that allowed us to estimate key background parameters such as the likely trait optima in the absence of the interaction (the "abiotic" optima).

The long-proboscid fly, M. longirostris, resides in lowland habitats near the coast of South Africa and pollinates a guild of at least 20 plant species [@manning1997themoegistorhynchus]. Among these species, the most widespread is L. anceps, a long-tubed perennial whose distribution extends outside the range of M. longirostris [@pauw2009flies]. Crucial to this analysis, I was able to estimate the likely optimal tube and proboscis lengths for these species in the absence of this particular interaction. Using the phenotypic data published by Pauw et al. [-@pauw2009flies], these parameters were inferred by averaging the trait values of additional species within the pollination assemblage.

The biotic selection strengths $B_1$ and $B_2$ on M. longirostris and L. anceps, respectively, do not statistically differ from zero ($p_1\approx$ 0.968, $p_2\approx$ 0.736, see Figure @networkfig). Thus, this analysis fails to support the hypothesis of strict pairwise coevolution in this system. The failure to find evidence of pairwise coevolution in this interaction may result from its actual absence or because the richness of the plant-pollinator community creates a scenario where coevolutionary selection is more diffuse and driven by a large number of weak pairwise interactions [@fox1988diffuse; @strauss2005toward]. In such a case, this approach may not be sufficiently powerful to detect the weak coevolution attributable to a single pairwise interaction. It is also possible that coevolution in this system is driven largely by indirect effects, undetectable with this approach, as has been recently suggested by [@guimaraes2017indirect].

```{r biostr, echo=F, fig.cap="(@biostr) \\ Inferred strengths of biotic selection", fig.align='center', anchor="figure"}
par(mar=c(0,0,0,0))
plot_jpeg('~/Documents/Website/bobweek.github.io/Bmtca.jpg',3)
```

I complemented this analysis of a plant-pollinator interaction with an analysis of the antagonistic interactions between the C. camellia weevil, an obligate seed predator of the C. japonica fruit [@toju2005imbalance]. Female weevils bore holes into the woody pericarps of the camellia to oviposit their offspring. Inside the fruit, weevil larvae feed on the seeds of the camellia up until the fourth instar, at which time they exit the fruit and overwinter [@toju2005imbalance]. These two species co-occur across Japan, although camellia populations where the weevil is absent also exist [@toju2005imbalance]. I estimated the strength of coevolution between the camellia and weevil using  data from previously published work [@toju2011metapopulation; @toju2011climatic] and the fact that male weevil rostrum lengths could be used as a proxy for the abiotic optimum of the female weevils since males do not interact with the camellia. The abiotic optima for the pericarp thickness of the camellia was inferred by averaging pericarp thicknesses across populations where weevils are absent. Both $B_1$, the strength of biotic selection on the weevil, and $B_2$, biotic selection on the camellia plant, significantly differed from zero ($p_1\approx$ 0, $p_2\approx$ 0, Figure (@biostr)). Hence, I found evidence for pairwise coevolution between the seed-eating weevil C. camellia and its host plant C. japonica. Unlike the plant-pollinator system, this interaction involves just two species. Hence there is no risk of diffusive coevolution masking the pairwise coevolutionary signal, making it easier to detect.

These results demonstrate that coupling existing coevolutionary models with a maximum likelihood approach allows the strength of coevolutionary selection to be estimated using routinely collected phenotypic data. By providing a methodology that does not rely on extensive and system specific experimental manipulation, this approach greatly expands the range of systems for which the strength of coevolutionary selection can be estimated, paving the road for a more quantitative and critical assessment of coevolution's importance in natural systems.  Hence, when coupled with data from a broad range of empirical systems, this method holds the potential to settle long standing debates involving the importance of species interactions and coevolution in the evolution of various phenomena including phenotypic diversity, sexual reproduction, community structure, and epidemiological dynamics [@yoder2010does; @hamilton1980sex; @nuismer2013coevolution; @anderson1982coevolution].

## Q3: What is the role of coevolution in structuring mutualistic communities?

There are several ways to describe ecological communities [@morin2009community]. One is taxonomic where the identities and abundances of species tend to be the primary focus. In this case the community can be broken down into sets of taxonomically related species called taxocenes. Another is functional where species are categorized by the ecological processes they engage in. These functional groups can be thought of as guilds, sets of species that partake in a similar use of resources, when (but not only when) the ecological processes that define the functional groups arise from inter-guild interactions centered around one guild attempting to procure resources from another. With this approach to describing a community of species, not only are the abundances of each species important for understanding the structure and dynamics of the community, but also the traits mediating the interactions and how both the traits and interactions evolve.

Here I use the latter description by focusing on interspecific interactions. Since a particular species may use different traits for each of the other species it interacts with, modelling the evolution of traits in a general community involving a wide variety of interaction types would require the development of multivariate models of trait evolution for each species across a mind-numbing number of possible interactions, giving rise to a highly complex model. But are the considerations of such immense detail necessary for the procurement of general principles underlying natural processes? This question elucidates two primary roles for the use of theory.

One use of theory is to tailor a model to the idiosyncratic details of a particular system in order to develop a quantitative understanding of the system. Using such an approach we may in principle predict the future state of a system or to infer events leading to the systems current state. For example, let us suppose that all of the species and interactions between them within a community have been fully described mathematically. Further suppose that accurate models of the evolution and demographic dynamics of each species have been derived from those descriptions. By tuning the parameters of the resulting models to reflect reality, we would be able to predict future population sizes and trait distributions, an enormous scientific achievement. However, this of course would require methods and data that do not exist at such a scale.

The second, which is the focus of the proposed work, is to discover basic principles that provide a qualitative understanding of the structure and behavior of the natural world. This approach maintains a more cognitive appeal by boiling down conclusions into statements that are made as simple as possible. By virtue of this goal, the associated mathematical models are developed to be as simple as possible. However, in order to derive such simple models from mechanistic considerations, a great deal of simplifying assumptions must be adopted. A reasonable question one should ask is “To what extent do the assumptions upon which a result has been obtained restrict the result from accurately capturing reality?” This question will be addressed for this project using simulations as will be discussed below.

The primary tool used for this project is theoretical quantitative-genetics. Hence, all of the assumptions of quantitative-genetic theory must be adopted. This includes assumptions such as the normality of phenotypic distributions, random mating, blending inheritance and constant additive genetic variance. Beyond the essential assumptions of quantitative genetic theory, I work with the two basic assumptions relevant at the community level: 1) The outcomes of interactions between any two members of separate guilds are mediated by the same traits in each species. This allows the theory to be developed around a single physical trait for each guild (such as proboscis or nectar tube lengths). 2) The mean phenotype and population size of any particular species are both treated as random variables. It will be assumed that for each species within a guild, these random variables are distributed independently and identically. This assumption transitions the models used from sets of equations describing the evolution and demographic dynamics of each species explicitly to a set of equations describing the dynamics of the distribution of traits and abundances of each guild as a whole. This last assumption is the essential component that makes this approach a mean-field approach. Starting with this set of assumptions, I will develop simple models of coevolving communities and use them to answer the following questions:

  * What distribution of trait values across a community emerges from biotic selection?
  * When does coevolution produce specialization?

This work takes the perspective that these two questions are intimately interwoven with each other. The coevolutionary process both effects and is affected by the network structure of the community as well as the distribution of functional traits across the community. As the network structure imposes dynamics on the traits mediating interactions, these changing traits affect the mode of interactions between species and therefore reciprocally impose dynamics on the network structure of the community. This observation is not particularly novel. Indeed, Nuismer et al. (2013) investigated various aspects of the relationship between coevolution and the network structure of mutualistic communities, finding that, among other results, strong matching coevolution tends to lead to higher modularity while weak matching coevolution results in nested communities.

The novelty of the proposed project comes from a goal to explicitly incorporate the behaviour of the organisms with respect to their interaction preferences. Previous models of coevolution focus on the role of functional trait values in mediating the outcomes of interactions. Here, in addition to predicting the effects of ecological interactions on fitness, the decision to interact, or the propensity of interactions to take place at all is incorporated into the model of coevolution. Assuming the interaction preference of an organism has a genetic basis, the evolution of this behavioral trait can be modelled. Using this approach I can develop models that determine the necessary and sufficient conditions for specialized interactions to evolve.

The empirical focus will be on pollination-based interactions and communities composed of such interactions. Therefore two guilds will form the basic structure of these communities, one consisting of the flower species and the other consisting of the pollinator species. Fittingly, there is a long-standing debate in the pollination literature on whether or not pollination as a process tends to produce higher degrees of specialization or instead more generalized interactions [@willmer2011pollination]. The argument for increased generalization comes from the idea that animal pollinators would profit from visiting any flower that benefits it. In turn, the various pollinators can act together as a selective force on floral traits to promote generalization in the flowers [@gomez2006ecological]. In contrast, it is thought that a plant would profit from having a pollinator that is specialized since this would increase the probability of conspecific pollen deposition [@willmer2011pollination]. Hence, the floral traits attracting more specialized pollinators should be selected for. By taking into account the merits of both of these arguments, it seems that pollination will not necessary act to monotonically drive a community towards either specialization or generalization, but depends upon certain details of the members of the community as well as their environment. This line of reasoning motivates the further development of theory to clarify the conditions under which pollination produces generalization or specialization or perhaps to maintain some intermediate state.

### Outline of the approach:

To quantify the degree of generalization of an individual I will consider the individuals interaction preference with respect to functional trait values (as opposed to taxonomic identity). This interaction preference can be represented by a statistical distribution across all possible trait values and is related in many ways to the notion of a niche [@vazquez2006community]. The particular value the individual chooses to interact with can be modelled using a random variable defined by the interaction preference distribution. For simplicity, I start by assuming this distribution is normal with mean $\kappa$ and variance $\gamma^2$. Then $\gamma>0$ can be understood as the degree of generalization of this individual. Hence, the distribution of $\bar\gamma$, the mean degree of generalization of a species, across each guild comprising the community provides a way to describe the network structure of the community that is similar to the information provided by the degree distribution, the number of nodes with a particular degree (see Figure $X$). If the community-level average $\bar\gamma$ is small (which implies that species are highly specialized on average) and has low variance (which implies that most species are highly specialized), we can anticipate the community to be highly modular. In contrast, if the community-level average $\bar\gamma$ is large (highly generalized species on average) and has low variance (most species are highly generalized), then we can anticipate the community to be highly nested (see Figure $X$). To model the evolution of $\gamma$ it can be treated as a continuous trait. However, because gamma is a positive number, we cannot assume it is normally distributed within a species as required by quantitative-genetic assumptions. Instead, by assuming $\gamma$ follows a log-normal distribution within any given species, the transformed variable $g=\ln\gamma$ will be normally distributed. This allows us to develop a model of the evolution of $g$, the mean-log degree of generalization for a given species, using the breeders equation. Extending this model to one that describes the evolution of the distribution of $g$ across the entire community will pave the way to theoretical answers for question 2 above.

→ Insert figures of degree distributions, distribution of generalization, network images for two guilds with high and low degrees of generalization (both with low variance). Use R to generate these images and describe how they were generated in the figure caption. ← 

For question 1, the focus is on functional trait evolution. Returning to the example of long-tongued flies and long-tubed flowers, the functional traits are the length of the tongues in the flies and the length of the tubes in the flowers. The coevolution of these traits has already been captured with the offset-matching model. However, that model contains the assumption of random encounters. Incorporating interaction preferences requires rederiving the model in a way that incorporates non-random encounters. Denoting $z$ the value of the functional trait (proboscis or tube length, for example), the resulting model describes the evolution of the guild-level mean and variance of the multivariate species-level mean trait $(\bar z,\bar κ,\bar g)$. This trait consists of one physical component, the mean functional trait of the species ($\bar z$), and two behavioral components, the mean preference of the species ($\bar\kappa$) and the mean generalization of the species ($\bar g$). For the sake of parsimony these traits are assumed to be genetically independent of one another.

To derive the model, we first write down the fitness function of an individual of species $i$ in guild 1 used to derive the offset-matching model

(@fitt1) $$W_{1,i}(x,y)=W_{1,\max}\exp\para{-\frac{(y+\delta_{1,i}-x)^2}{2\beta_{1,i}}}$$

Here $x$ denotes the functional trait of an individual in guild 1 (say the proboscis length of a pollinator) while $y$ denotes the functional trait of an individual in guild 2 (such as the nectar tube length of a plant). For this project, focus will be restricted to just two guilds, consistent with the focus on pollination communities.

To obtain the mean fitness of this individual we must consider the distribution of the trait value, $Y$, this individual will interact with. As a random variable, this trait value that will be drawn from all of the species of the adjacent guild. Hence we need to calculate the distribution of the random variable $Y$. Denote this distribution by $p_2(y)$. Likewise, the distribution of the functional trait for guild 1 will be denoted by $p_1(x)$

The next step is to incorporate the interaction preference of the individual. The density of this distribution is denoted $\iota_1(y)$ and as implied above, is expressed as

(@iota) $$\iota_1(y)=\frac{1}{\sqrt{2\pi\gamma_1^2}}\exp\para{-\frac{(y-\kappa_1)^2}{2\gamma_1^2}}.$$

Thus, the mean fitness of an individual in guild 1 of species $i$ with trait $x$ is

(@fitt2) $$W_{1,i}(x)=\int W_{1,i}(x,y)\phi_2(y)\iota_1(y)\eta_2(y)dy.$$

Finally, the mean fitness of species i in guild 1 can be written as

(@meanfitt) $$\overline W_{1,i}=\int W_{1,i}(x)\phi_{1,i}(x)dx$$

Hence, the change in $\bar x_i$ is [@lande1976natural]

(@traitdyn) $$\Delta\bar x_i=G_{1,i}\frac{\partial\ln\overline{W}_{1,i}}{\partial\bar x_i}.$$


<!-- As a cognitive crutch, assume for now that there are $K$ species in guild 2 with abundances $N_1, N_2, \dots, N_K$. Then, assuming each species’ functional trait is distributed normally with mean $\bar y_i$ and variance $\sigma^2_{2,i}$, we have

(@y) $$Y\sim \mathcal{N}\para{\sum P_{2,i}\bar y_i, \sum P_i^2\sigma^2_{2,i}}$$

where $P_i=\frac{N_i}{\sum N_i}$ is the proportion of the adjacent guild composed of species $i$. At this point we can allow the $N_i$ to be random variables as well, derive the distribution of the $P_i$ and find the resulting distribution of $Y$. Instead, as a first pass, it is reasonable to approximate this calculation by using the average $P_i$, denoted $\bar p$, in place of each $P_i$. Of course $\bar p$ is simply calculated as

(@meanp) $$\bar p=\frac{1}{K}\sum P_i=\frac{1}{K}.$$

Hence, we obtain

(@y2) $$Y\sim \mathcal{N}\para{\frac{1}{K}\sum\bar y_i, \frac{1}{K^2}\sum\sigma^2_i}.$$

By remaining consistent with the quantitative genetic assumption of constant phenotypic variance we have that $\bar\sigma^2\equiv\frac{1}{K^2}\sigma^2_i$ is a constant.

The next step is to incorporate the interaction preference of the individual. The density of this distribution is denoted $\iota_1(y)$ and as implied above, is expressed as

(@iota) $$\iota_1(y)=\frac{1}{\sqrt{2\pi\gamma_1^2}}\exp\para{-\frac{(y-\kappa_1)^2}{2\gamma_1^2}}.$$

Hence, the mean fitness of an individual in guild 1 of species $i$ with trait $x$ is

(@fitt2) $$W_{1,i}(x)=\int W_{1,i}(x,y)\phi_2(y)\iota_1(y)dy=C\exp\para{-\frac{\gamma_1^2(\bar y+\delta_1-x)^2+(\sigma_2^2+\beta_1^2)(\bar y-\kappa_1)^2}{2(\sigma_2^2\gamma_1^2+\sigma_2^2\beta_1^2+\gamma_1^2\beta_1^2)}}.$$

$$\approx C\exp\para{-\frac{}{}}, \ \beta_1^2\gg\sigma_2^2,\gamma_1^2$$

Recall, however, that both $\kappa$ and $\gamma$ are random variables. So the above equation is not quite right. Instead it should be written

(@fitt3) $$W_{1,i}(x)=\int\int\int W(x,y)\phi(y)\iota(y)dy \ f_1(\kappa)d\kappa \ h_1(\gamma)d\gamma.$$

where $f_1(\kappa)$ and $h_1(\gamma)$ are the densities associated with the distributions of $\kappa$ and $\gamma$ across guild 1 respectively.

Clearly we run into an issue with gamma and instead of finding the formal mathematical resolution to this problem, we will charge forward by approximating the random variable gamma with its mean gamma_bar. Finally, the mean fitness of species i in guild 1 can be written as

(@meanfitt) $$\overline W_{1,i}=\int W_{1,i}(x)\phi_{1,i}(x)dx=\dots$$

Hence, the change in $\bar x_i$ is

(@traitdyn) $$\Delta\bar x_i=\dots.$$

-->

By extending this model to capture the evolution of the two behavioral traits, $\kappa_i$ and $\gamma_i$ (or $g_i=\ln\gamma_i$ to be more precise), as well, expressions describing the community-wide means and variances of these three traits in each guild can be calculated. In particular, denoting $\mu_i=(\E \bar z_i,\E \bar κ_i,\E \bar g_i)^\top$ and $\nu_i=\para{\V\bar z_i,\V\bar κ_i, \V\bar g_i}^\top$ to be the multivariate mean vector and vector of variances of the three traits across guild $i$, it would be ideal to find relatively simple equations describing the evolution of these quantities of the form

(@commmean) $$Δμ_i=U_{1,i} μ_i+V_{1,i} \nu_i$$

(@commvar) $$\Delta\nu_i=U_{2,i} \mu+V_{2,i} \nu_i$$

where the $U_i$ and $V_i$ are 3x3 matrices.

The parameters anticipated to be most important in determining when specialization will occur are the strengths of biotic selection ($B$) and the optimal offset ($δ$). By imposing specific statistical distributions on these parameters across the community, it may be discovered that, for example, a high variance in $δ$ inevitably leads to the specialization in interactions between a subset of the community.

Although biotic selection is an important component of coevolution, it alone does not represent a coevolutionary process. Coevolution is defined by the reciprocal response of two interacting populations to selection induced by their interaction. Denoting $B_1$ as the random variable representing the strength of biotic selection induced on guild 1 by guild 2 and $B_2$ the random variable representing strength of selection on guild 2 by guild 1, then $\sqrt{B_1B_2}$, the geometric mean of the two biotic selection strengths is a natural way to quantify the potential for coevolution between the two guilds. Thus, the distribution of this geometric mean across interacting pairs will capture the relevance of coevolution throughout the community. Relating the properties of this distribution, such as its mean and variance, to specific ecological and evolutionary outcomes, such as the evolution of specialization, will provide important insight into the many ways that coevolution shapes ecological communities.

This analytical approach will also be complemented by a set of genetically-explicit individual-based simulations to probe the less tractable situations that arise when assumptions are broken. For example, if the distribution of population sizes across the community is allowed to fluctuate according to the mean fitnesses of those populations (already defined by the evolutionary equations required to derive the coevolutionary model), it is not clear whether or not the resulting model will be keen for traditional mathematical analysis. Hence, for situations such as this one, finely tuned high resolution simulations of this process may provide more insight than raw mathematical analysis.

## References