---
output:
  html_document:
    css: style.css
    theme: cosmo
    toc: yes
    toc_float: yes
title: "Proposed thesis: The coevolution of mutualistic communities"
author: "Bob Week"
csl: evolution.csl
bibliography: ref.bib
---

```{r setup, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)
```

```{r results="asis", echo=F}
cat("
<style>
a {
   color: black;
}
</style>
")
```

\newcommand{\W}{\mathcal{W}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\para}[1]{\left( #1 \right)}
\newcommand{\sqbr}[1]{\left[ #1 \right]}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Cov}{\mathrm{Cov}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\Matrix}[1]{\left(\begin{matrix}#1\end{matrix}\right)}

<!--
## Summary

The coevolutionary process has been speculated to have played a significant role throughout the history of life ever since the time of Darwin. Previous studies have used coevolution to provide putative explanations for the evolution of sexual reproduction, the structuring of ecological communities, the extreme exaggeration observed in many physical traits and for the diversity of life.

This
-->

## Introduction to proposed work

Coevolution, the reciprocal evolutionary response of two populations to their interactrion, has long been thought to play a key role in driving the exaggeration of traits [@darwin1859origin;@ehrlich1964butterflies], promoting major evolutionary transitions such as the evolution of sexual reproduction [@hamilton1980sex;@jaenike1978hypothesis;@lively1987evidence;@gandon2007evolution], manipulating the structure of ecological communities [@nuismer2013coevolution], influencing epidemiological dynamics [@anderson1982coevolution;@lion2015evolution;@lively2016coevolutionary], and maintaining the diversity of life [@yoder2010does]. However, despite coevolution's long suspected importance, a rigorous understanding of both the mechanisms involved and the patterns produced by this process is still in its infancy. A clear understanding of the role that coevolution plays in the evolutionary play must account for the details of the mechanisms driving the coevolutionary process, be able to quantify its strength and significance in nature and determine its consequences for ecological communities. The proposed research seeks to further develop past theoretical work contributing to these three main objectives.

A major limiting factor in past theoretical work comes from a common modelling approach. Previous models capturing the relationship between community ecology and coevolution have focused on results pertaining to a small number of interacting species (cite roughgarden, abrams, holt and mcpeek for theory, strauss, benkman, berenbaum, zergall for empirical work). The common approach in each of these studies is to derive a model of mean trait evolution alongside a model of demographic dynamics for each species involved. Hence, if there are $N$ species under consideration, there will be $2N$ dynamical, often nonlinear, equations used to model the system. This approach has provided some analytical results for a few species, but rapidly becomes intractable with additional species. However, many of the patterns observed in ecological communities span dozens, even hundreds of species [@morin2009community;@hubbell2011unified;@manning1997themoegistorhynchus]. To overcome this difficulty a recent development in coevolutionary theory has taken the so-called mean field approach with respect to the identity of species (cite paper that's in review and scott's 2013 paper and a book or article on the mean-field approach). This approach allows the reduction of a high-dimensional problem involving equations describing the evolution of each species within a community to a low-dimensional one that focuses on the community as a whole. With this approach analytical results can be derived that capture the statistical nature of the community, but without explicit conclusions for any particular member of the community. Hence, the stage has been set to develop a mathematically rigorous theory of coevolutionary community ecology. The ultimate goal of the proposed work is to contribute to this development through extending previous work on the coevolution of mutualistic networks by incorporating further biological elements than have been previously considered (such as behaviour and demographic dynamics, see section Q3).

However, before such a contribution can be made there are two major steps to be taken. First, a more biologically realistic treatment of the mechanisms mediating the outcomes of ecological interactions needs to be incorporated into coevolutionary theory. Variation in this mechanism alone may explain a significant amount of functional diversity seen in nature. Second, a metric for the strength of coevolution within a community is required to quantify the role of coevolution in the community. A way to measure the strength of coevolution between a pair of species will be needed in order to derive this community-wide metric from first principles. Hence, a novel metric of pairwise-coevolution and a method to infer it will be introduced. Accomplishing these first two steps will brandish a framework ripe for rigorous investigations of the relationship between coevolution and ecological communities. 
To tether this work to the empirical world, each step is motivated by the details of specific pollination networks. The primary example used throughout is the interaction between a long-tongued fly Moegistorhynchus longirostris and the flower it pollinates, Lapeirousia anceps, occuring in the coastal lowlands of South Africa [@pauw2009flies;@manning1997themoegistorhynchus]. Other fly-flower mutualisms from across the globe will be explored for cross-system comparisons. 

<!-- However, the end goal is to find a well-studied mutualistic network that is large and taxonomically rich from which methods can be developed to tie theory to data. -->

The three primary questions driving this work are:

  1. Why do some mutualistic pairs produce highly exaggerated traits while others do not?
  2. What is the strength of pairwise coevolution in the wild?
  3. What is the role of coevolution in structuring mutualistic communities?

These questions are complemented by the following three overarching goals:

  1. To develop a model that explicitly accounts for the mechanism mediating mutualistic interactions.
  2. To develop a method for inferring the strength of coevolution between a pair of species.
  3. To develop models of coevolving mutualistic communities that track the evolution of specialization.


## Q1: Why do some mutualisms produce highly exaggerated traits while others do not?

#### (Introduction to the offset-matching model of coevolution)

Mutualistic interactions provide an abundant supply of putative coevolutionary arms races. The most popular of these being the potentially coevolved orchid Angraecum sesquipedale and the moth that pollinates it Xanthopan morganii praedicta. The extreme length of this orchids nectar tube had captured the attention of Charles Darwin [-@darwinvarious]. Darwin formulated the hypothesis that this exaggerated trait may be the result of the reciprocal evolutionary responses between this orchid and whatever had been pollinating it (unknown to Darwin at the time) over many generations. In a rebuttle to a criticism of Darwins idea, Alfred Russel Wallace provided some logic to defend Darwin [@arditti2012good]. Wallace had been carefully thinking about the mechanism mediating the outcome of this interaction. He speculated that the orchid would be better off having a nectar tube that is slightly longer than the proboscis of the pollinator, forcing the pollinator to further penetrate the flower in order to obtain its nectar reward. Wallace speculated that this would cause the body of the pollinator to rub against the reproductive organs of the flower and thereby increase the probability of pollination. Hence, it can expected that selection induced by this interaction will select for slightly longer tube lengths. Likewise, with respect to the fitness of the pollinator, Wallace put forward that by having a proboscis slightly longer than the tube length of the flower, the pollinator would be able to extract more resources from each flower it visits. Thus, selection from this interaction would be reciprocally selecting for individuals from both populations with slightly longer traits. Assuming that these traits maintain ample genetic variation and that the primary source of selection on these traits is induced by their interaction, then, ignoring other processes such as migration and drift, the logical result is a never ending arms race towards more and more exaggerated, yet closely matched, traits.

Given the logic of Wallaces argument, one might expect any pollination-based mutualism mediated by a mechanism analogous to the one described above to consistently produce highly exaggerated traits. However, a wide variety in the degree of exaggeration among traits mediating such interactions across a variety of plant-pollinator interactions has been observed [@anderson2010predictable]. What can account for such variance? Clearly there is an endless list of potential explanations. However, by focusing on explanations pitched entirely in terms of the structure of the interactions between pairs of species, a clear understanding of the role that coevolution plays in the maintenance of diversity can be developed.

<!-- Do I need to incorporate more of the anderson paper? -->

Since different species are often placed under different physiological constraints and environmental conditions, it seems obvious that this variation should exists across species. However, even within species, a striking amount of variation in the outcomes of these processes has been observed. Take, for example, the interaction between the fly Moegistorhynchus longirostris and the flower it pollinates, Lapeirousia anceps [@pauw2009flies]. The mechanism mediating this interaction is precisely the same as that described above for A. sesquipedale and X. morganii. The mean traits of the fly and flower species covary spatially with a correlation coefficient of 0.78, seeming to confirm the prediction that these traits should be closely matched. Yet, across eight populations the average length of the fly's proboscis ranges from 43 mm to 86 mm and the average nectar tube of the flower across the same populations from 41 mm to 77 mm. Then, given that each population is undergoing the same interaction, why is there such stark phenotypic variability across these species' distributions? To answer this question I propose to use a novel quantitive genetic model of coevolution derived from considering the mechanism mediating pollination interactions.

### Outline of the approach:

Previous work on the coevolutionary theory of quantitative traits have focused on two primary mechanisms; trait matching and trait differences. In the case of trait matching, it is assumed that the optimal phenotype of each species is the mean phenotype of the complementary species. This model results in very well behaved linear dynamics described by

(@match)
$$\Delta\bar z_i=G_iB_i(\bar z_j-\bar z_i).$$

Here, $G_i$ denotes the additive genetic variance of the trait $z_i$, the small non-negative parameter $1\gg B_i\geq0$ is the strength of biotic selection induced by species $j$ onto species $i$ and $\bar z_i,\bar z_j$ represent the mean traits in species $i$ and $j$ respectively. This model predicts the mean traits for each species to converge to a common value that lies somewhere between their initial phenotypes. Hence, no escalation can occur in this model. 

In contrast, @nuismer2010correlation developed a model based on trait differences that leads to pure escalation dynamics. The resulting dynamics of this model are

(@diff)
$$\Delta\bar z_i=G_i\sqrt{B_i}$$

Under this model the mean traits of each species will forever increase at a constant rate independently of one another. Though this model predicts an arms race, it misses a crucial ingredient described by Darwin, that the traits in each species need only be slightly longer than the traits of the complementary species to maximize fitness. 

By incorporating this notion of an optimal offset into the trait matching model, I have coalesced these two distinct sets of dynamics just described within a unifying framework that I call the offset-matching model. This generalized model of coevolution takes the form

(@coal)
$$\Delta\bar z_i=G_iB_i\delta_i+G_iB_i(\bar z_j-\bar z_i)$$

where the additional term $\delta_i$ represents the offset between the traits of species $i$ and $j$ that maximizes the fitness benefits of species $i$ with repsect to their interaction. By setting $\delta_i=0$ for both species I obtain the classic trait matching model. Since $B_i$ is small it can be shown that by setting $\delta_i=1/\sqrt{B_i}$ the offset-matching model is approximated by the pure escalation model when $|\bar z_j-\bar z_i|$ is small.

Using this model of offset-matching I can explain both the escalation and spatial covariation observed in nature. I plan to use the offset-matching model, focusing on mechanism of the interaction as manifested by the offset parameter $\delta$, as a platform to understand why there is such variation in the outcomes of these processes across the spatial distribution of coevolving pairs. Assuming constant strengths of biotic selection in two interacting species throughout their populations, two explanations for the variability observed include the stochastic effects of drift and spatial variability in the offset term. These two explanations are not mutually exclusive and each taken on their own may require biologically unrealistic parameters in order to mimic real data. Hence, it is likely that a combination of random drift and spatial variability in mechanism will be able to explain the patterns observed within a biologically relevent subset of parameter space.

The key parameters to be investigated are the variance of drift $(V_d)$ and the variance in offset $(V_\delta)$ for each species. In order to couple this theory to the empirical world I plan to use an evolutionary algorithm to find values for $(V_d,V_\delta)$ that produces simulated data whose summary statistics roughly match those of the data published in Pauw et al. [-@pauw2009flies]. For this algorithm to work, a population of values for the pair $(V_d,V_\delta)$ is provided upon which a subset is selected for "reproduction." The criteria for selection will involve the comparison of summary statistics calculated from data simulated under these parameter pairs to the summary statistics drawn from the empirical system. Reproduction involves imperfect replication of the selected pairs, allowing the algorithm to search parameter space. This approach is not as formal as maximum likelihood or Bayesian approaches. However, its advantages include its relative ease of implementation and computational tractability.

Though the quantitative genetic approach has the strength of providing analytic results, its weakness lies in the large amount of assumptions required to obtain such results. To account for this restriction, I plan to test the depency of my conclusions on the assumptions made by developing individual-based simulations that allow me to numerically derive results for generalized models. By comparing analytical results to simulations, I can quantify the extent to which particular assumptions prevent accurate descriptions of biological processes. Specifically, statistical tests, such as the t-test, can be used to calculate whether or not simulated results significantly differ from those predicted by quantitative genetic theory.

## Q2: What is the strength of coevolution in the wild?

In spite of the much anticipated importance of the coevolutionary process [@thompson1994coevolutionary], almost nothing is known about its the strength and prevalence in the wild. This lack of knowledge is largely due to the absence of generally applicable statistical approaches that yield quantitative estimates for the strength and significance of coevolution. To fill this critical gap I propose a novel statistical method of inference that derives maximum likelihood estimates for the strength of pairwise coevolution in natural systems by coupling the offset-matching model of coevolution to widely available phenotypic data. I will apply this method to two well-studied interactions; one being a highly specific antagonism involving the Japanese camellia fruit and its predatory weevil and the other being the mutualism previously discussed between the long-tongued fly M. longirostris and the flower it pollinates L. anceps. Broad application of this approach has the potential to answer long-standing evolutionary debates such as the importance species interactions play in the evolution of sexual reproduction [@hamilton1980sex].

Current understanding of coevolution's importance rests upon methods that fall into two general classes: those that are broadly applicable but yield only qualitative evidence for coevolution and those that produce quantitative estimates for the strength of coevolution but can be applied only in a narrow range of systems. For example, one popular approach for inferring coevolution relies on measuring the spatial correlation between traits of interacting species and using significant interspecific correlations as evidence of a coevolutionary process [@berenbaum1986constraints;@benkman2003reciprocal;@hanifin2008phenotypic;@toju2008fine;@pauw2009flies]. Strengths of this approach include the relative ease of collecting the relevant data and its broad applicability to a wide range of species interactions. The critical weakness of this approach, however, is that significant interspecific correlations are neither necessary nor sufficient for demonstrating coevolution  [@nuismer2010correlation;@janzen1980coevolution] and can never yield more than qualitative evidence for coevolution. In contrast, quantitative approaches for estimating the strength of coevolution, such as selective source analysis, require the collection of extensive trait and fitness data from interacting species and thus have proven difficult to employ in all but a few specialized systems [@ridenhour2005identification;@brodie2003reciprocal;@sletvold2010pollinator;@toju2005imbalance]. As a consequence of these trade-offs in existing approaches, rigorous, quantitative estimates of the strength of coevolution in natural populations are extremely scarce. 

### Outline of the approach:

A promising alternative to existing approaches is the development of a likelihood-based framework that uses routinely collected phenotypic data to estimate key parameters of the offset-matching model of coevolution and to test for the existence of coevolution. Specifically, the offset-matching model can be easily extended to predict the spatial distribution of population-level mean traits for a pair of interacting species that evolve in response to random genetic drift, abiotic selection, and coevolution [@nuismer2010correlation]. Crucially, this extension predicts that the distribution of trait means in the interacting species across a metapopulation will approach a bivariate normal distribution entirely described by five statistical moments: the metapopulation-level average of the population-level average key trait in each species among populations ($\mu_1$ and $\mu_2$), the variance of the population-level mean trait in each species among populations ($V_1$ and $V_2$), and the spatial association (covariance, $C$) between the key traits in each species. The key model parameters that quantify the strength of biotic selection acting on the two species are $B_1$ and $B_2$. Any selective agents acting independently of the interaction are captured collectively by the strengths of abiotic selection on each species; $A_1$ and $A_2$. All four of these model parameters are required to be non-negative and small ($\ll1$). Denoting the optimal phenotype with respect to abiotic selection by $\theta_i$ and the effective population size by $n_i$ for species $i$, classical quantitative genetic assumptions lead to the following recursion equations for the statistical moments describing the metapopulation:

(@dynmet)
$$\mu_i'=\mu_i+G_i\sqbr{B_i\delta_i+B_i(\mu_j-\mu_i)+A_i(\theta_i-\mu_i)}$$
$$V_i'=(1-2A_iG_i)V_i+2B_jG_j(C-V_i)+G_i/n_i$$
$$C'=(1-A_1G_1-A_2G_2)C+B_2G_2(V_1-C)+B_1G_1(V_2-C).$$

Because the models predict a bivariate normal distribution of traits, calculating the likelihood of observing any particular set of trait values in a pair of interacting species is straightforward. In particular, letting $D=\{D_i\}_{i=1}^N\equiv\{(\bar z_1,\bar z_2)_i^\top\}_{i=1}^N$ denote the data (mean trait pairs of the interacting species from several populations) one obtains

(@lik)
$$L(D|\mu,\Sigma)=\prod_{i=1}^N\frac{\exp\para{-\frac{1}{2}(D_i-\mu)^\top\Sigma^{-1}(D_i-\mu)}}{2\pi\sqrt{|\Sigma|}}, \ \Sigma=\Matrix{V_1 & C\\ C & V_2}.$$

Using the equilibrium of system (@dynmet) expressions of the first five moments can be derived in terms of the selection strengths $A_1,A_2,B_1$ and $B_2$. Unfortunately these functions cannot be inverted to obtain the selection strengths as functions of the first five moments. Thus, I numerically solve for the $A_i$ and $B_i$ that maximize 

(@likstr)
$$L(D|A_1,A_2,B_1,B_2)\equiv L(D|\mu(A_1,A_2,B_1,B_2),\Sigma(A_1,A_2,B_1,B_2)).$$

By maximizing this likelihood with respect to the strengths of biotic and abiotic selection acting on the two species a quantitative estimate for the strength of coevolution is obtained under the assumption that the remaining model parameters have been accurately estimated.

In addition to providing quantitative estimates for the strength of biotic selection, this method can be used to rigorously test for the presence of coevolution. Specifically, for a coevolutionary hypothesis to be supported, reciprocal selection must be demonstrated [@janzen1980coevolution;@thompson1994coevolutionary]. In this maximum likelihood framework, this long-standing and widely accepted definition of coevolution corresponds to demonstrating that the pair of parameters in this model that quantify the strength of biotic selection imposed on each species by the other are significantly greater than zero. By performing likelihood ratio tests, support for the coevolutionary hypothesis can be compared relative to support for the non-coevolutionary null hypotheses of unilateral evolution where $B_1=0$ or $B_2=0$, or evolution entirely independent of species interactions where $B_1=0$ and $B_2=0$ (Figure @networkfig). Due to the nested structure of these models, I can directly compare the likelihood of coevolution against the likelihoods of these null models by approximating p-values. Figure @networkfig shows that both $p_1$ and $p_2$ need to be less than the significance threshold $\alpha$ to support a coevolutionary hypothesis.

```{r, include=F}
require('jpeg')
plot_jpeg = function(path, scale, add=FALSE)
{
  jpg = readJPEG(path, native=T) # read the file
  res = scale*dim(jpg)[2:1] # get the resolution, [x, y]
  if (!add) # initialize an empty plot area if add==FALSE
    plot(1,1,xlim=c(1,res[1]),ylim=c(1,res[2]),asp=1,type='n',xaxs='i',yaxs='i',xaxt='n',yaxt='n',xlab='',ylab='',bty='n')
  rasterImage(jpg,1,1,res[1],res[2])
}
```

```{r networkfig, echo=F, fig.cap="(@networkfig) \\ Relationships between the coevolutionary hypothesis and non-coevolutionary hypotheses", fig.align='center', anchor="figure"}
plot_jpeg('~/Documents/Website/bobweek.github.io/network.jpg',2)
```

Restricting the selection parameters $B_i$ to match the null hypotheses of unilateral evolution ($B_1=0$ or $B_2=0$) results in a restricted likelihood ($L(D|A_1,A_2,0,B_2)$ or $L(D|A_1,A_2,B_1,0)$), one that is always less than $L(D|A_1,A_2,B_1,B_2)$. Then calculating the significance of $B_1$, for example, is equivalent to computing the probability that the statistic

(@likrat)
$$\Lambda=2\ln\para{\frac{L(D|A_1,A_2,B_1,B_2)}{L(D|A_1,A_2,0,B_2)}}$$

lands within a given percentile of its distribution under the assumption of the null hypothesis $B_1=0$. I adopt the standard 95th percentile as the threshold for determining significance.

According to Wilks' Theorem, the distribution of this statistic is approximated by a Chi-square for large sample sizes. However, typical numbers of populations sampled are on the order of ten. Hence, this approximation fails and empirical distributions need to be calculated  [@wasserman2013all]. To do so, 250 sets of data under the null model are simulated for each comparison, with each set of data containing the number of populations equal to those sampled empirically. This provides adequate resolution (p-value estimates span 0 to 1 in increments of 0.004) while remaining computationally feasible. Calculations have been carried out in the statistical programming language $R$ and scripts will be made available at:

https://github.com/bobweek/inferring.coevolution


### Current Results:

I applied this maximum likelihood approach to two textbook examples of coevolution where previous work has implicated coevolution as a cause of trait exaggeration and spatial variability [@pauw2009flies; @toju2011weevils]: the interaction between the long-tongued fly M. longirostris and a plant it pollinates L. anceps and the interaction between the camellia plant Camellia japonica and its seed predator, the weevil Curculio camellia. In both cases, the interactions are thought to depend largely on a single key trait in each species (fly proboscis and plant floral tube lengths or weevil rostrum length and camellia pericarp thickness). In addition to the essential phenotypic data, previous work in both systems provided valuable additional information that allowed us to estimate key background parameters such as the likely trait optima in the absence of the interaction (the "abiotic" optima).

The long-proboscid fly, M. longirostris, resides in lowland habitats near the coast of South Africa and pollinates a guild of at least 20 plant species [@manning1997themoegistorhynchus]. Among these species, the most widespread is L. anceps, a long-tubed perennial whose distribution extends outside the range of M. longirostris [@pauw2009flies]. Crucial to this analysis, I was able to estimate the likely optimal tube and proboscis lengths for these species in the absence of this particular interaction. Using the phenotypic data published by Pauw et al. [-@pauw2009flies], I inferred these parameters by averaging the trait values of additional species within the pollination assemblage. Details of the methods used to calculate other background parameters can be found in the supplementary information (SI $\S$C). I found that the biotic selection strengths $B_1$ and $B_2$ on M. longirostris and L. anceps, respectively, do not statistically differ from zero ($p_1\approx$ 0.968, $p_2\approx$ 0.736, see Figure @networkfig). Thus, this analysis fails to support the hypothesis of strict pairwise coevolution in this system. The failure to find evidence of pairwise coevolution in this interaction may result from its actual absence or because the richness of the plant-pollinator community creates a scenario where coevolutionary selection is more diffuse and driven by a large number of weak pairwise interactions [@fox1988diffuse; @strauss2005toward]. In such a case, this approach may not be sufficiently powerful to detect the weak coevolution attributable to a single pairwise interaction. It is also possible that coevolution in this system is driven largely by indirect effects, undetectable with this approach, as has been recently suggested by [@guimaraes2017indirect].

```{r biostr, echo=F, fig.cap="(@biostr) \\ Inferred strengths of biotic selection", fig.align='center', anchor="figure"}
plot_jpeg('~/Documents/Website/bobweek.github.io/Bmtca.jpg',3)
```

I complemented this analysis of a plant-pollinator interaction with an analysis of the antagonistic interactions between the C. camellia weevil, an obligate seed predator of the C. japonica fruit [@toju2005imbalance]. Female weevils bore holes into the woody pericarps of the camellia to oviposit their offspring. Inside the fruit, weevil larvae feed on the seeds of the camellia up until the fourth instar, at which time they exit the fruit and overwinter [@toju2005imbalance]. These two species co-occur across Japan, although camellia populations where the weevil is absent also exist [@toju2005imbalance]. I estimated the strength of coevolution between the camellia and weevil using  data from previously published work [@toju2011metapopulation; @toju2011climatic] and the fact that male weevil rostrum lengths could be used as a proxy for the abiotic optimum of the female weevils since males do not interact with the camellia. The abiotic optima for the pericarp thickness of the camellia was inferred by averaging pericarp thicknesses across populations where weevils are absent. Both $B_1$, the strength of biotic selection on the weevil, and $B_2$, biotic selection on the camellia plant, significantly differed from zero ($p_1\approx$ 0, $p_2\approx$ 0, Figure @biostr). Hence, I found evidence for pairwise coevolution between the seed-eating weevil C. camellia and its host plant C. japonica. Unlike the plant-pollinator system, this interaction involves just two species. Hence there is no risk of diffusive coevolution masking the pairwise coevolutionary signal, making it easier to detect.

These results demonstrate that coupling existing coevolutionary models with a maximum likelihood approach allows the strength of coevolutionary selection to be estimated using routinely collected phenotypic data. By providing a methodology that does not rely on extensive and system specific experimental manipulation, this approach greatly expands the range of systems for which the strength of coevolutionary selection can be estimated, paving the road for a more quantitative and critical assessment of coevolution's importance in natural systems.  Hence, when coupled with data from a broad range of empirical systems, this method holds the potential to settle long standing debates involving the importance of species interactions and coevolution in the evolution of various phenomena including phenotypic diversity, sexual reproduction, community structure, and epidemiological dynamics [@yoder2010does; @hamilton1980sex; @nuismer2013coevolution; @anderson1982coevolution].

This project was submitted on December 7th, 2017 as a letter to the International journal of science, __Nature__, and is currently in review.

## Q3: What is the role of coevolution in structuring mutualistic communities?

There are several ways to describe ecological communities [@morin2009community]. One is taxonomic where the identities and abundances of species tend to be the primary focus. In this case the community can be broken down into sets of taxonomically related species called taxocenes. Another is functional where species are categorized by the ecological processes they engage in. These functional groups can be thought as guilds, sets of species that partake in a similar use of resources, when (but not only when) the ecological processes that define the functional groups arise from inter-guild interactions centered around one guild attempting to procure resources from another. With this approach to describing a community of species, not only are the abundances of each species important for understanding the structure and dynamics of the community, but also the traits mediating the interactions and how both the traits and interactions evolve.

Here I use the latter description by focusing on interspecific interactions. Since a particular species may use different traits for each of the other species it interacts with, modelling the evolution of traits in a general community involving a wide variety of interaction types would require the development of multivariate models of trait evolution for each species across a mind-numbing quantity of interactions, giving rise to a highly complex network structure. It is undoubtable that if such a framework were successfully constructed in a biologically realistic yet mathematically tractable manner, it would provide not only deep insight into the manifold relations between coevolution and community ecology but also an improved ability to predict the response of ecological communities to changing climates and the movement of species. However, before taking on such a grandiose and difficult task, there are obvious steps to take that result in models which are more humble and tractable but still contribute to the development of this grand vision. Beyond the essential assumptions of quantitative genetic theory, such as normally distributed phenotypes, the three basic assumptions I work with are: 1) The outcomes of interactions between any two members of separate guilds are mediated by the same traits in each species. This reduces the multivariate models required above to univariate ones. 2) The distribution of species abundances across the community remains fixed. This assumption can be taken a step further by assuming this distribution is uniform and to ignore species abundances altogether. 3) I assume that each species within a guild is distributed indepedently and identically with respect to both trait value and distribution. This assumption transitions the models used from sets of equations describing the evolution and demographic dynamics of each species explicitly to a set of equations describing the dynamics of the distribution of traits of each guild as a whole. This last assumption is the essential component that makes this approach a mean-field approach. Starting with this set of assumptions, I will develop simple models of coevolving communities and use them to answer the following questions:

  1. What predicts the distribution of trait values across a community?
  2. When does coevolution produce specialization?

This work takes the perspective that these two questions are intimately interwoven with each other. The coevolutionary process both effects and is affected by the network structure of the community as well as the distribution of functional traits across the community. As the network structure imposes dynamics on the traits mediating interactions, these changing traits affect the mode of interactions between species and therefore reciprocally impose dynamics on the network structure of the community. This observation is not particularly novel. Indeed, Nuismer et al. [-@nuismer2013coevolution] investigated various aspects of the relationship between coevolution and the network structure of mutualistic communities, finding that, among other results, strong matching coevolution tends to lead to higher modularity while weak matching coevolution results in nested communities. 

The novel contribution of the proposed project comes from a goal to explicitly incorporate the behaviour of the organisms with respect to their interaction preferences. Previous models of coevolution focus on the role of functional trait values in mediating the outcomes of interactions. Here, in addition to predicting the effects of ecological interactions on fitness, I model the decision, or the propensity of interactions to take place at all. Assuming the interaction preference of an organism has a genetic basis I can model the evolution of this behavioral trait. Using this approach I can develop models that determine the necessary and sufficient conditions for specialized interactions to evolve.

The focus will be on pollination-based interactions and communities composed of such interactions. Therefore two guilds will form the basic structure of these communities, one consisting of the flower species and the other consisting of the pollinator species. Fittingly, there is a long-standing debate in the pollination literature on whether or not pollination as a process tends to produce higher degrees of specialization or instead more generalized interactions [@willmer2011pollination]. The argument for increased generalization comes from the idea that animal pollinators would profit from visiting any flower that benefits it. In turn, the various pollinators can act together as a selective force on floral traits to promote generalization in the flowers [@gomez2006ecological]. In contrast, it is thought that a plant would profit from having a pollinator that is specialized since this would increase the probability of conspecific pollen deposition [@willmer2011pollination]. Hence, the floral traits attracting more specialized pollinators should be selected for. By taking into account the merits of both of these arguments, it seems that pollination will not necessary act to monotonically drive a community towards either specialization or generalization, but depends upon certain details of the members of the community as well as their environment. I am therefore interested in further developing theory to clarify the conditions under which pollination produces generalization or specialization or instead acts to maintain some intermediate state.

### Outline of the approach:

To quantify the degree of generalization of an individual I will consider the individuals interaction preference with respect to functional trait values (as opposed to taxonomic identity). This interaction preference can be represented by a statistical distribution across all possible trait values and is related in many ways to the notion of a niche [@vazquez2006community]. I can model the particular value the individual chooses to interact with using a random variable described by the interaction preference distribution. For simplicity, I start by assuming this distribution is normal with mean $\kappa$ and variance $\gamma^2$. Then $\gamma>0$ can be understood as the degree of generalization of this individual. Hence, the distribution of $\gamma$ across the community provides a way to describe the network structure of the community that is similar to the information provided by the degree distribution. If the average $\gamma$ is small and has low variance, I can anticipate the community to be highly modular. In contrast, if the average $\gamma$ is large and has low variance, then I can anticipate the community to be highly nested. To model the evolution of $\gamma$ I can treat it as a continuous trait. Assuming $\gamma$ follows a log-normal distribution within the population of any particular species, the transformed variable $g=\ln\gamma$ will be normally distributed. This allows us to develop a model of the evolution of $\bar g$, the mean-log degree of generalization for a given species, using the breeders equation. Extending this model to one that describes the evolution of the distribution of $\bar g$ across the entire community will pave the way to theoretical answers for question 2 above.

For question 1, the focus is on functional trait evolution. Returning to the example of long-tongued flies and long-tubed flowers, the functional traits are the length of the tongues in the flies and the length of the tubes in the flowers. The coevolution of these traits has already been captured with the offset-matching model. However, that model contains the assumption of random encounters. If I are to incorporate interaction preferences then the model will need to be rederived in a way that incorporates non-random encounters. Denoting $z$ the value of the functional trait (proboscis or tube length, for example), the resulting model will be describing the evolution of the multivariate mean trait $(\bar z, \bar\kappa, \bar g)$. For the sake of parsimony I assume these traits to be genetically independent of one another. Extending this model to the whole community, expressions describing the community-wide means, variances and covariances of these traits can be calculated. In particular, denoting $\mu=\E(\bar z, \bar\kappa, \bar g)$ and $\Sigma=\Var(\bar z, \bar\kappa, \bar g)$ to be the multivariate mean vector and covariance matrix of the three traits across the community, I hope to find relatively simple equations describing the evolution of these quantities of the form:

$$\Delta\mu=f_1(\mu,\Sigma)$$
$$\Delta \Sigma=f_2(\mu,\Sigma).$$

The parameters I anticipate to be most important in determining when specialization will occur are the strengths of biotic selection ($B$) and the optimal offset ($\delta$). By imposing specific statistical distributions on these parameters across the community, it may be discovered that, for example, a high variance in $\delta$ inevitably leads to the specialization in interactions between a subset of the community. 

Although biotic selection is an important component of coevolution, it alone does not represent a coevolutionary process. Coevolution is defined by the reciprocal induction of biotic selection between two interacting populations. Denoting $B_f$ as the random variable representing the strength of biotic selection induced on the flowers by the pollinators and $B_p$ the random variable representing strength of selection on the pollinators by the flowers, then $\sqrt{B_fB_p}$, the geometric mean of the two biotic selection strengths is a natural way to quantify the strength of coevolution between the two guilds. Thus, the distribution of this geometric mean across interacting pairs will capture the presence of coevolution throughout the community. Relating the properties of this distribution to specific ecological and evolutionary outcomes, such as the evolution of specialization, will provide important insight into the many ways that coevolution shapes ecological communities.

This analytical approach will also be complemented by a set of simulations that probe the less tractable situations that arise when assumptions are broken. For example, if the distribtuion of population sizes across the community is allowed to fluctuate according to the mean fitnesses of those populations (already defined by the evolutionary equations required to derive the coevolutionary model), it is not clear whether or not the resulting model will be keen for analysis. Hence, for situations such as this one, finely tuned high resolution simulations of this process may provide more insight than raw mathematical analysis.

## Side projects

Should I elaborate on any of these?

  * With Scott
      * Coevolution slows the dissassembly of communities 
      * Coevolution of reaction norms
      * Duration of coevolutionary associations
      * Non-random encounters
  * Community phylogenetics with Megan
  * Evolving diversification rates with Austin
  * Demographic dynamics of coevolving communities
      * Invasion dynamics
      * $\alpha/\beta$-diversity
      * Speciation
      * Community assembly

## References