\documentclass{resume}
\usepackage{layouts}

% title / author / date
\name{Generic Name}
\address{New York, NY}


% macro for highlighting code

% for vrb
% for tables
% for graphics

\usepackage{multicol}
\setlength{\multicolsep}{2.0pt plus 1.0pt minus 0.5pt}% 50% of original values

\begin{document}


\begin{rSection}{Education}
            {\bf Hudson University} \hfill {2007--2009} \\
                        GPA 3.955\\
        Master of Arts in Statistics
    \smallskip\\
            {\bf Smallville College} \hfill {2000--2004} \\
                        Cumulative GPA 3.838 Summa Cum Laude, Honors in Mathematics\\
        Bachelor of Science in Mathematics, Journalism Minor\\
        The Wayne Award for Excellence in Mathematics\\
        Member of Pi Mu Epsilon, a national honorary mathematics society
    \end{rSection}

\begin{rSection}{Highlights}
\begin{multicols}{2}
\setlength\multicolsep{0pt}
Author of \emph{Pulitzer Prize} winning article \\
Organizer of \textbf{Glasses and Cowl} Meetup \\
Analyzed global survey by the \textbf{Surveyors Inc} \\
Professor of Journalism at \textbf{Hudson University} \\
Thesis on \textbf{Facial Recognition Errors} \\
Served as reporter in \textbf{Vientiane, Laos}
\end{multicols}
\end{rSection}

\begin{rSection}{Relevant Experience}

\begin{rSubsection}{Pied Piper}{2013--Present}{CTO}{New York, NY}
\item Set up company's computing platform
\item Constructed statistical models for predictive analytics of big data
\end{rSubsection}

\begin{rSubsection}{Goliath National Bank}{2011--2013}{Quant}{New York, NY}
\item Built quantitative models for derivatives trades
\item Wrote algorithms using the R statistical programming language
\end{rSubsection}

\begin{rSubsection}{Goliath National Bank}{2010}{Intern}{New York, NY}
\item Got coffee for senior staff
\item Killed time by pretending I was using Excel after finishing all work in R
\item Hung out on Stone Street drinking beer and eating pizza
\end{rSubsection}

\begin{rSubsection}{Surveyors Inc}{2008--2010}{Data Scientist}{Chicago, IL}
\item Analyzed survery data disimilar to the Census
\item Ported Stata code to R code
\end{rSubsection}

\begin{rSubsection}{Daily Planet}{2006--2008}{Reporter}{New York, NY}
\item Partnered with a Pulitzer Prize winning writer and a cub reporter
\item Answered directly to the Editor-in-Chief
\item Used this position to identify world events that needed mending
\end{rSubsection}

\begin{rSubsection}{Hudson University}{2011--Present}{Professor}{New York, NY}
\item Taught class on using data in journalism
\item Debunked myths about facial recognition in press photos
\item Built syllabus to encourage collaboration
\end{rSubsection}

\begin{rSubsection}{Hooli}{2004--2006}{Programmer}{San Fransisco, CA}
\item Worked on compression software projects
\item Attained near-impossible Weissman Score for 3D compression
\item Experience paved the path for Pied Piper years later
\end{rSubsection}

\end{rSection}

\begin{rSection}{Athletics}
\textbf{Ice Hockey} \emph{Goaltender} | \textbf{Hudson University} | 2000--2004 \\
\textbf{Curling} \emph{Vice Skip} | \textbf{Hudson University} | 2000--2004
\end{rSection}

\clearpage

\begin{rSection}{Research Experience}

\begin{research}{Hudson University}{Humanitarian Research}{2008--Present}
To reduce the time, cost and effort necessary to properly analyze a region affected by a natural or manmade disaster Dr. Richard Oddie and I worked on methods to show that smaller sample sizes can adequately provide the same information as larger sample sizes.  For our study we used data from the 2010 Pakistan floods that devastated large swathes of the country.  We compared varying samples sizes ranging from 5 tehsils (villages) in each province to full enumeration.  While the research is still ongoing we have so far demonstrated that as few as 10 tehsils per province is sufficient.  This should help improve future responses to disasters.


I was engaged in statistical and analytic support to Dr. Oddie of Hudson's School of Nursing and Dr. John Nilla from Hudson's World Institute in their study of the socioeconomic, health and geographic factors affecting deaths in New Orleans following Hurricane Katrina. We are reconciling a self-reporting death list with an official State list that is thought to under represent Katrina related deaths. The goal is to determine what disparities, if any, resulted in a greater proportion of deaths among certain populations as opposed to others. The analysis will result in journal article preparation in in which I will be one of the authors.


\end{research}

\begin{research}{Daily Planet}{Does the World Need Superman?}{2008}
Over the course of two years wrote an in-depth expose examining the need for Superman.  While I did not draw any conclusions it led many people to question their beholdence to a supernatural being from another planet.  Ultimately, however, most people agreed that Superman was indeed essential to the planet, even as just a beacon of hope.


\end{research}

\end{rSection}


\end{document}